### 2020-07-03 1-Main.tex
Initial Overleaf Import

\documentclass[sigchi]{acmart}

\usepackage[colorinlisttodo]{todonotes}

\usepackage{listings}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

\DeclareTextFontCommand{\mytexttt}{\ttfamily\hyphenchar\font=45\relax}

\copyrightyear{2020}
\acmYear{2020}
\setcopyright{acmlicensed}
\acmConference[Workshop XXX '2020]{A Workshop on Software Engineering}{2020}{Somewhere}
\acmBooktitle{Workshop XXX '2020: Workshop on Software Engineering, Somewhere}
%\acmPrice{15.00}
%\acmDOI{10.1145/1122445.1122456}
%\acmISBN{978-1-4503-9999-9/18/06}

\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{PAMELA: an annotation-based Java Modelling Framework}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Sylvain Guérin, Guillaume Polet, Caine Silva, Joel Champeau}
%\orcid{1234-5678-9012}
\affiliation{%
  \institution{Lab-STICC \\ ENSTA Bretagne}
%  \streetaddress{P.O. Box 1212}
  \city{Brest}
  \country{France}
  %  \postcode{43017-6221}
}
\email{firstname.lastname@ensta-bretagne.fr}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{S. Guérin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.

\input{2-Abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Model-Driven Engineering, Java Modelling Framework, Meta-modeling, Models@Runtime}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\input{3-Introduction}

\input{4-ApproachOverview}

\input{5-Features}

\input{6-Experimentations}

\input{7-Relatedworks}

\input{8-Conclusion}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}

\end{document}
\endinput
%%
%% End of file `sample-sigchi.tex'.

### 2020-07-03 2-Abstract.tex
Initial Overleaf Import

\begin{abstract}
This article presents PAMELA, an annotation-based Java Modelling framework. PAMELA provides a smooth integration between model and code, and enable Java developers to handle software development both at conceptual level and at source-code level, without code transformation and/or generation, and avoiding round-tripping issues.

% TODO...

% Parler aussi de Aspect Oriented Programming, Traits programming, Run-time weaving, Semantics Add-On, MOP (Meta Object Protocol)

\end{abstract}

### 2020-07-03 4-ApproachOverview.tex
Initial Overleaf Import

\section{Approach overview}

We present in this section an overview for PAMELA approach.

\subsection{Model serialization in source code}

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. The idea is to avoid separation between modeling and code to facilitate consistency management and avoid round-tripping issues.

To do so, we argue that source code is the right artefact to encode the model with metadata information stored in tagged code. This requires an annotation-enabled language. Such language supports the attribute-oriented programming if its grammar allows adding custom declarative tags to annotate standard program elements. Java programming language from version 1.5 is a good candidate with the support of annotations.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{PamelaVision.pdf}
    \caption{PAMELA approach for modelling}
    \label{fig:PamelaVision}
\end{figure}

Figure \ref{fig:PamelaVision} shows PAMELA approach for storing model in source code. The model is inlined in many source code files, with a set of annotations covering PAMELA metamodel as presented in the next subsection.

\subsection{PAMELA use process}

Coupling model and code into the same artefact open new ways of programming. The classical way relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provides the right annotations the programmers use. This is, for instance, the process followed by JEE developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer becomes also the modeler. In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel. Our experience shows that, in that case, introducing and reusing new concepts (1) reduce the size of the code, (2) reduce the risk of errors and (3) improve the code structure. The cycle of development between the model and the code can then be drastically reduced, leading to what we call \emph{continuous modeling}.

The code size is reduced because abstractions factorize the recognize pattern and the previous code is replaced by the use of the abstraction at the right place. This also reduce the risk since the previous code is now generated by the PAMELA framework with all the required verification. And finally, the code structure is improved since it matches the way the programmer conceptualizes (models) her/his code.

The developed metamodels are implemented by annotations that relies on Java/JVM entities and mechanisms. They include consistency checking that constrains their use and help the programmer. We have experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or roles and rules to set security rules on classes.

\subsection{PAMELA metamodel}

PAMELA metamodel is presented in figure \ref{fig:PamelaMetaModel}.
This metamodel is classical and reflects a common class diagram vision such as in UML.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{PamelaMetaModel.pdf}
    \caption{PAMELA metamodel}
    \label{fig:PamelaMetaModel}
\end{figure}

\subsubsection{PAMELAModel}

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}.

\subsubsection{ModelEntity}

A \texttt{ModelEntity} reflects a concept and is encoded in a java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. A \texttt{ModelEntity} also defines some properties, encoded as \texttt{ModelProperty}. Note that reification of \texttt{ModelEntity} is performed in a java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods. A partially implemented \texttt{abstract} java texttt{class} may be defined as partial base implementation (conform to implemented interface).

\subsubsection{ModelProperty}

A \texttt{ModelProperty} is identified by a name, a cardinality (simple or multiple) and a type, which can be a reference to another \texttt{ModelEntity}, or a Java type (a primitive or an arbitrary complex Java type). Depending on its cardinality, a \texttt{ModelProperty} is bound to a set of methods reflecting use of property.
\begin{itemize}
    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a java method defined in java interface taking no argument and returning desired value).
    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a java method taking value to be set as unique argument)
    \item A \emph{read-write multiple property} will define a \emph{getter}, a \emph{adder} (a java method taking value to be added as unique argument), a \emph{remover} (a java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
\end{itemize}

A strong interest of the approach is that the model is encoded in java, and must be compiled. It forces the java compiler to perform required checks for a PAMELA model encoded in a strong typed program. Execution semantics of model is fully compatible with Java semantics. Many validation rules are automatically performed through classical java compilation, independently of underlying PAMELA execution semantics.

\subsection{A basic example}

The following code listing represents a very basic model with two entities \emph{Book} and \emph{Library}. Entity \emph{Book} defines two read-write single properties \emph{title} and \emph{ISBN} with single cardinality and with \texttt{String} type. Entity \emph{Book} also define a constructor with initial \emph{title} value. Entity \emph{Library} defines a read-write multiple properties \emph{books} referencing \emph{Book} instances. Note that this code is sufficient to execute the model, while no line of code is required (only java interface and API methods are declared here).

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize]
@ModelEntity
public interface Book 
  extends AccessibleProxyObject {
  @Initializer
  public Book init(@Parameter("title") 
    String aTitle);
  @Getter("title")
  public String getTitle();
  @Setter("title")
  public void setTitle(String aTitle);
  @Getter("ISBN")
  public String getISBN();
  @Setter("ISBN")
  public void setISBN(String value);
}

@ModelEntity
public interface Library
  extends AccessibleProxyObject {
  @Getter(value = "books", 
    cardinality = Cardinality.LIST)
  public List<Book> getBooks();
  @Adder("books")
  public void addToBooks(Book aBook);
  @Remover("books")
  public void removeFromBooks(Book aBook);
  @Reindexer("books")
  public void moveBookToIndex(Book aBook, 
    int index);
  @Finder(collection = "books",
    attribute = "title")
  public Book getBook(String title);
}
\end{lstlisting}
 %   \caption{A basic PAMELA model with two entities \texttt{Library} and \texttt{Book}}
 %   \label{fig:ABasicPamelaModel}
%\end{figure}

Execution of this model may be performed using following simple lines of code.

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize]
// Instantiate the meta-model
// by computing the closure of concepts graph
ModelContext modelContext = ModelContextLibrary
  .getModelContext(Library.class);
// Instantiate the factory
ModelFactory factory 
  = new ModelFactory(modelContext);
// Instantiate a Library
Library myLibrary 
  = factory.newInstance(Library.class);
myLibrary.setName("My library");
// Instantiate some Books
Book myFirstBook = factory.newInstance(
  Book.class, "Lord of the ring");
Book anOtherBook = factory.newInstance(
  Book.class, "Holy bible");
myLibrary.addToBooks(myFirstBook);
myLibrary.addToBooks(anOtherBook);
\end{lstlisting}
 %   \caption{Executing a PAMELA model}
 %   \label{fig:ExecutingPamelaModel}
%\end{figure}

The first line of code instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships. This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled byte-code. A factory \texttt{ModelFactory} is then instantiated using that \texttt{ModelContext}, allowing to instantiate \emph{Library} and \emph{Book} instances.

\subsection{Handling custom code}

A major challenge to be addressed by MDE (Model-Driven Engineering) approaches is the ability to integrate custom implementation to a base of code derived from a formal model. The major drawback is the way back (round-tripping).

PAMELA frameworks provides an elegant way to do it, while using common extension points such as inheritance, as offered by java language. Custom implementations should be declared in java classes, this or those classes declared as partial implementation(s) of related \emph{ModelEntity}.

The following example shows how to integrate custom code to the fully interpreted \emph{Book} entity described above. The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in annotation header of model entity. Custom implementations are defined using classical java implementation/overrides scheme. Here we define implementation of \texttt{read()} method, which has no annotation (and thus unable to be processed by PAMELA framework), and also implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property. Note that this implementation references default interpretated implementation (call to \texttt{performSuperGetter(String)} method).

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize]
@ModelEntity
@ImplementationClass(BookImpl.class)
public interface Book 
  extends AccessibleProxyObject {
  static final String TITLE = "title";
  @Getter(TITLE)
  String getTitle();
  // ... title property declarations ...
  void read();
}
// Provides a partial implementation for Book
public static abstract class BookImpl 
  implements Book {
  @Override
  public String getTitle() {
    String title = performSuperGetter(TITLE);
    if (title == null) {
      return "This book has no title";
    }
    return title;
  }
  @Override
    public void read() {
      // do the job
    }
}
\end{lstlisting}
 %   \caption{Executing a PAMELA model}
 %   \label{fig:ExecutingPamelaModel}
%\end{figure}

As said previously, PAMELA framework supports multiple inheritance. In this context, it is possible to provide multiple partial classes as implementation for a given \emph{ModelEntity}. To do so, we use abstract inner classes tagged with \texttt{@Implementation}, and the composition is made at run-time.

\subsection{Run-time considerations}

Resulting model execution is a combination of:
\begin{itemize}
    \item plain java byte-code, as the result of the basic compilation of source code,
    \item and an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty} declarations.
\end{itemize}

This composition offers many benefits: 
\begin{itemize}
    \item Strong coupling between model and code
    \item Strong typing is kept, and required checks are performed by the java compiler
    \item PAMELA framework provides interpretation of model@runtime
    \item No need to generate POJO (plain old java objects), as their execution follow the standard semantics (less code, less bugs)
    \item Custom implementation are provided if needed, using classical java extension points
    \item It offers a way to intercept method calls and instrument the code
\end{itemize}

From a technical point of view, PAMELA implementation uses \emph{javassist} reflection library, providing \texttt{MethodHandler} mechanism, which is a way to override the java dynamic binding. Invoking a method on an object which is part of a PAMELA model, caused the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed. This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at run-time (support for Design by Contract, aka JML).
% TODO: trouver des références pour ca

PAMELA framework is a 100\% pure java (> 1.5), compilable by a classical java compiler and executable in a classical Java Virtual Machine.

% Rajouter un paragraphe sur le fait que PAMELA est une implementation du MOP (meta object protocol) ?

\subsection{Common PAMELA annotations}

Here is a non-exhaustive list of the most common java annotations reflecting PAMELA metamodel.

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declares an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, enventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique reindexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in infered metamodel while \texttt{ModelContext} computation.
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances

\end{itemize}

### 2020-07-03 5-Features.tex
Initial Overleaf Import

\section{PAMELA features}

This section presents an overview of major features available in the context of PAMELA framework use.

\subsection{Meta-model at run-time}

Metamodel closure is computed on runtime, working on the classpath of launched java application, and starting from a simple java interface (or a collection of java interfaces) which is/are PAMELA-annotated. From a mathematical point of view, internal representation of the metamodel is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}. \mytexttt{@Imports} and \mytexttt{@Import} annotations allows to include some other \emph{ModelEntities} in the metamodel. On the contrary, an annotation attribute \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that context, metamodel computation is a graph closure computation, starting from a collection of vertices.

Metamodel closure computation on-the-fly provides an interesting approach to deal with model fragmentation.

% TODO: references

\subsection{PAMELA objects life-cycle management}

A Metamodel computation is represented by a \mytexttt{ModelContext} and uses a \mytexttt{ModelFactory} built with that model context to handle instances of that metamodel. The \mytexttt{ModelFactory} is responsible of the life-cycle of instances of metamodel (construction and destruction of Java instances). \mytexttt{@Initializer} annotation allows to define parametered constructors for \emph{ModelEntity} instances.

\mytexttt{AccessibleProxyObject} is a Java interface providing utilities methods which are interpreted by internal PAMELA interpreter. This includes calls to internal code execution, such as \mytexttt{performSuperSetter(String,Object)} which represent a call to internal setter of property identified by supplied String value.

Figure \ref{fig:LifeCycle} illustrates life-cycle of objects beeing instantiated as PAMELA instances. The \mytexttt{ModelFactory} initiates creation and triggers right constructor during a phase where the object is in \mytexttt{isCreating} status.

\begin{figure}
    \centering
    \includegraphics[width=0.9 \columnwidth]{LifeCycle.pdf}
    \caption{Life-cycle of PAMELA objects}
    \label{fig:LifeCycle}
\end{figure}

Modifications of objects are internally tracked by PAMELA interpreter which manages \mytexttt{Modified} status, according to containment semantics as presented further (a contained object modification implies object flagged as modified, and implied container flaggued as modified too). Saving object graph brings back object status in \mytexttt{Alive/Saved} status.

Since calls to any features of model objects are dispatched by the internal interpreter, PAMELA runtime offers a multi-level undo/redo stack tooling. When enabled, this scheme allows to store and manage an edition model composed of atomic edits. Figure \ref{fig:AtomicEditMetaModel} presents atomic edits metamodel for a fine-grained model modification tracking system. That mechanism provides undo/redo features, virtually unlimited. For performance reasons, we can set a maximum depth for undo/redo operations.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{AtomicEditMetaModel.pdf}
    \caption{Atomic edits metamodel}
    \label{fig:AtomicEditMetaModel}
\end{figure}

\mytexttt{DeletableProxyObject} provides delete (and undelete) features to \emph{ModelEntity} instances. Deletion features are performed using a context which is a graph closure computation for all the objects which have to be deleted. PAMELA also offers undelete feature which allow to resurrect a deleted object. Deleted objects are still resurectable until they are still in the undo/redo stack. A deleted object who is leaving scope of maximum depth of undo/redo stack is destroyed. Object is fully dereferenced, ready for garbage collecting, and cannot be resurrected again.

\subsection{Meta-programming support}

Figure \ref{fig:PamelaMetaModel} represents PAMELA metamodel. Concept of \emph{ModelProperty} represent access to a value (simple or with multiple cardinality). This read-access is implicitely implemented using \mytexttt{get:} protocol. Following the same logic, PAMELA properties may expose following protocols:
\begin{itemize}
    \item \mytexttt{get:} read-access to a data. This protocol is implemented by all kind of \emph{ModelProperties}.
    \item \mytexttt{set:} write-access to a data. This protocol is implemented by \emph{ModelProperties} implementing \mytexttt{SettablePropertyImplementation} API.
    \item \mytexttt{add:} and \mytexttt{add:AtIndex:} add-access to a multiple cardinality data. These protocols are implemented by \emph{ModelProperties} implementing \mytexttt{MultiplePropertyImplementation} API.
    \item \mytexttt{remove:} remove-access to a multiple cardinality data. This protocols is implemented by \emph{ModelProperties} implementing \mytexttt{MultiplePropertyImplementation} API.
    \item \mytexttt{delete:} and \mytexttt{undelete:} delete/undelete protocols, implemented by all \emph{ModelProperties}
\end{itemize}

Unless specified in \mytexttt{@PropertyImplementation} PAMELA annotations, default implementation are provided the PAMELA interpreter. Default implementations are encoded in well identified classes which are extendable.

PAMELA allows programmers to define their custom implementation for a given set of properties, while providing a class implementing \mytexttt{@PropertyImplementation} API and some protocol implementation (depending on the nature of the \emph{ModelProperty}. Following excerpt of code illustrate use of two custom property implementations.

\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize]
@ModelEntity
public interface MyConcept {

	static final String VALUE = "value";
	static final String SUB_CONCEPTS = "someSubConcepts";

	@Getter(value = VALUE)
	@PropertyImplementation(MyPropertyImplementation.class)
	String getValue();

	@Setter(VALUE)
	public void setValue(String value);

	@Getter(value = SUB_CONCEPTS, cardinality = Cardinality.LIST)
	@PropertyImplementation(MyListCardinalityPropertyImplementation.class)
	@Embedded
	List<MySubConcept> getSubConcepts();

	@Adder(SUB_CONCEPTS)
	void addToSubConcepts(MySubConcept subConcept);

	@Remover(SUB_CONCEPTS)
	void removeFromSubConcepts(MySubConcept subConcept);
}

public class MyPropertyImplementation extends DefaultSinglePropertyImplementation<Concept, String> {

	public MyPropertyImplementation(ProxyMethodHandler<Concept> handler, ModelProperty<Concept> property)
			throws InvalidDataException {
		super(handler, property);
	}

    // Implements get: protocol
    // ('concatenate' semantics)
	@Override
	public void set(String aValue) throws ModelDefinitionException {
		String oldValue = get();
		if (get() != null) {
			super.set(aValue + get());
		}
		else {
			super.set(aValue);
		}
	}
}

public class MyListCardinalityPropertyImplementation<I, T> extends AbstractPropertyImplementation<I, List<T>>
		implements MultiplePropertyImplementation<I, T> {
	...
	// Implements get: protocol
	public List<T> get() {
		...
	}
	// Implements add: protocol
	public void addTo(T aValue) throws ModelDefinitionException {
		...
	}
	// Implements remove: protocol
	public void removeFrom(T aValue) {
		...
	}
}
\end{lstlisting}

Such mechanisms are really usefull to control and fine-tune code implementation. Suppose for example that you have a code used in monothreading context: all your multiple cardinality properties will rely for example on a \mytexttt{ArrayList} implementation. Instead of replacing all occurences of \mytexttt{ArrayList} by \mytexttt{Vector} (multi-thread safe version of Java \mytexttt{List}) in generated code, the developer has just one modication to do, in adequate \mytexttt{PropertyImplementation}.

\subsection{Multiple inheritance and traits programming}

\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize]
@ModelEntity
public interface IntegerStorage extends AccessibleProxyObject {

     public static final String STORED_VALUE = "storedValue";

     @Getter(value = STORED_VALUE, defaultValue = "-1")
     public int getStoredValue();

     @Setter(value = STORED_VALUE)
     public void setStoredValue(int aValue);

     public void reset();

     @Implementation
     public abstract class IntegerStorageImpl implements IntegerStorage {
	@Override
	public void reset() {
	     setStoredValue(0);
	}

	@Override
	public void setStoredValue(int aValue) {
	     performSuperSetter(STORED_VALUE, aValue);
	     System.out.println("Sets stored value to be " + aValue);
	}
    }
}

@ModelEntity
public interface PlusProcessor extends IntegerStorage {

     public void processPlus(int value);

     @Implementation
     public abstract class PlusProcessorImpl implements PlusProcessor {
	@Override
	public void processPlus (int value) {
	     setStoredValue(getStoredValue()+value);
	     return getStoredValue();
	}
     }
}

@ModelEntity
public interface MinusProcessor extends IntegerStorage {

     public void processMinus(int value);

     @Implementation
     public abstract class MinusProcessorImpl implements MinusProcessor {
	@Override
	public void processMinus (int value) {
	     setStoredValue(getStoredValue()-value);
	     return getStoredValue();
	}
     }
}

@ModelEntity
public interface Calculator extends PlusProcessor, MinusProcessor {

}

\end{lstlisting}

\subsection{Embedding management and cloning support}

- embedding support
- cloning support (closure computation)
- clipboard operations

\subsection{Notification support}

- notification support
- isModified() support (embedding semantics)

\subsection{Persistance support}

- XML serialization/deserialization

\subsection{Graph computation features}

- support for computing equality
- diff/merge support, updateWith()
- visiting support

\subsection{Support for Contrat Programming}

- design by contract
- contract programming (JML)

% reference to jContractor

\subsection{Design patterns / Aspect programming / Code weaving}

- design patterns / aspect programming / code weaving

### 2020-07-03 6-Experimentations.tex
Initial Overleaf Import

\section{PAMELA Experimentation}

https://pamela.openflexo.org
https://github.com/openflexo-team/pamela

% Parler des différentes expériences faites avec PAMELA

\subsection{Performance issues}

% parler des problèmes de perfs

\subsection{Syntaxic issues}

% Problème d'un truc trop verbeux > editeur graphique

### 2020-07-03 3-Introduction.tex
Test

\section{Introduction}

This is a test.

Model-Driven Software Development focuses on managing domain-specific models, which represent concerns at conceptual level. Those models are generally represented as various artefacts, in many languages and representations. Model-Driven approach also generally requires code generators, which allow to generate source code. Underlying semantics of code to be executed in generally encoded in those code generators and can be inlined or implicitly defined by code generation process, or may be sometimes explicit to the code generation. Figure \ref{fig:ClassicalVision} shows the classical vision for for Model-Driven Engineering.

A first major drawback of this approach is the gap created between the conceptual level (the model) and the source code, where semantics may be totally hidden or implicit. Another classical issue resides in the development or maintenance process, where model and source code must evolve independently (and sometimes by different actors, eg architect and developer). Round-tripping mechanisms are commonly used to overcome those difficulties, but those tools are difficult to use and maintain, and the process may lead to inextricable issues in a context of concurrent modifications of model  and source code. For those reasons, major software developments projects abandon the idea of maintaining the link between model and source code during development process. In that context, the model is developed at the early stages of development process, and is used to prototype software application. The model may be manually reviewed back at the end of development process, for documentation purposes.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{ClassicalVision.pdf}
    \caption{Classical vision for Model-Driven Engineering}
    \label{fig:ClassicalVision}
\end{figure}

% Challenges description

This paper’s contributions are: ...

% Contributions

This paper is structured as follows: The second section focuses on ....

% Plan de l'article

### 2020-07-03 4-ApproachOverview.tex
Ajout paragraphe PAMELA use process

@@
 In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.
-Our experience shows that, in that case, introducing and reusing new concepts (1) reduce the size of the code, (2) reduce the risk of errors and (3) improve the code structure.
-The cycle of development between the model and the code can then be drastically reduced, leading to what we call \emph{continuous modeling}.
-
-The code size is reduced because abstractions factorize the recognize pattern and the previous code is replaced by the use of the abstraction at the right place.
-This also reduce the risk since the previous code is now generated by the PAMELA framework with all the required verification.
-And finally, the code structure is improved since it matches the way the programmer conceptualizes (models) her/his code.
-
 The developed metamodels are implemented by annotations that relies on Java/JVM entities and mechanisms.
@@
 We have experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or roles and rules to set security rules on classes.
+
+Our experience shows that introducing and reusing new concepts (1) reduce the size of the code, (2) reduce the risk of errors and (3) improve the code structure.
+The cycle of development between the model and the code can then be drastically reduced, leading to what we call \emph{continuous modeling}.
+
+The code size is reduced because abstractions factorized the recognize pattern and the previous code is replaced by the use of the abstraction at the right place.
+This also reduces the risk since the previous code is now generated by the PAMELA framework with all the required checks.
+And finally, the code structure is improved since it matches the way the programmer conceptualizes (models) her/his code.
 

### 2021-01-27 1-Main.tex
Test
fix usepackage error
Started rebuttal process

@@
-\documentclass[sigchi]{acmart}
+%\documentclass[preprint]{acmart}
+\documentclass[preprint]{elsarticle}
+\journal{Science of Computer Programming}
 
-\usepackage[colorinlisttodo]{todonotes}
+%\usepackage[colorinlisttodo]{todonotes}
 
@@
 \usepackage{xcolor}
+\usepackage{hyperref}
+\usepackage{tikz-uml}
 
@@
     showtabs=false,                  
-    tabsize=2
+    tabsize=2https://www.overleaf.com/project/5efee0e8ad6bc10001d47854
 }
@@
 
-\copyrightyear{2020}
-\acmYear{2020}
-\setcopyright{acmlicensed}
-\acmConference[Workshop XXX '2020]{A Workshop on Software Engineering}{2020}{Somewhere}
-\acmBooktitle{Workshop XXX '2020: Workshop on Software Engineering, Somewhere}
 %\acmPrice{15.00}
@@
 \begin{document}
+\begin{frontmatter}
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{PAMELA: an annotation-based Java Modelling Framework}
+\title{PAMELA: an annotation-based Java Modeling Framework}
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-\author{Sylvain Guérin, Guillaume Polet, Caine Silva, Joel Champeau}
-%\orcid{1234-5678-9012}
-\affiliation{%
-  \institution{Lab-STICC \\ ENSTA Bretagne}
-%  \streetaddress{P.O. Box 1212}
-  \city{Brest}
-  \country{France}
-  %  \postcode{43017-6221}
-}
-\email{firstname.lastname@ensta-bretagne.fr}
+%% Group authors per affiliation:
+%\author{Sylvain Guérin, Guillaume Polet, Caine Silva, Joel Champeau}
+%\address{ENSTA Bretagne}
 
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-\renewcommand{\shortauthors}{S. Guérin, et al.}
+%\author{Jean-Christophe Bach, Salvador Martinez, Antoine Beugnard}
+%\address{IMT Atlantique}
 
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
+%% or include affiliations in footnotes:
+\author[mymainaddress]{Sylvain Guérin\corref{mycorrespondingauthor}}
+\cortext[mycorrespondingauthor]{Corresponding author}
+%\ead{firstname.lastname@ensta-bretagne.fr}
+
+\author[mymainaddress]{Guillaume Polet}
+%\ead{firstname.lastname@ensta-bretagne.fr}
+
+\author[mymainaddress]{Caine Silva}
+%\ead{firstname.lastname@ensta-bretagne.fr}
+
+\author[mymainaddress]{Joel Champeau}
+%\ead{firstname.lastname@ensta-bretagne.fr}
+
+\author[mysecondaryaddress]{Jean-Christophe Bach}
+%\ead{firstname.lastname@imt-atlantique.fr}
+
+\author[mysecondaryaddress]{Salvador Mart\'inez}
+%\ead{firstname.lastname@imt-atlantique.fr}
+
+\author[mysecondaryaddress]{Antoine Beugnard}
+%\ead{firstname.lastname@imt-atlantique.fr}
+
+\address[mymainaddress]{ENSTA Bretagne, Lab-STICC, UMR 6285, Brest, France}%vérifier si c'est l'affiliation officielle
+\address[mysecondaryaddress]{IMT Atlantique, Lab-STICC, UMR 6285, F-29238 Brest, France}
 
@@
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
+\begin{keyword}
+Model Driven Engineering \sep Object Oriented Programming \sep Metaprogramming 
+\end{keyword}
 
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Model-Driven Engineering, Java Modelling Framework, Meta-modeling, Models@Runtime}
+\end{frontmatter}
 
@@
 %% information and builds the first part of the formatted document.
-\maketitle
 
-\input{3-Introduction}
+\section{Introduction}
+\label{sec:introduction}
+\input{introduction.tex}
 
-\input{4-ApproachOverview}
+\section{Approach}
+\label{sec:pamproc}
+\input{pamelaProcess.tex}
 
-\input{5-Features}
+\section{Software Framework and Features}
+\label{sec:architecture}
+\input{architecture.tex}
+
+\section{Implementation}
+\label{sec:implementation}
+\input{implementation.tex}
+
+%\section{Industrial experimentation and validation}
+%\label{sec:validation}
+%\input{validation.tex}
 
 \input{6-Experimentations}
+\label{sec:Experimentations}
 
-\input{7-Relatedworks}
+\section{Related Work}
+\label{sec:related}
+\input{related.tex}
+
+%\input{4-ApproachOverview}
+
+%\input{5-Features}
+
+%\input{6-Experimentations}
+
+%\input{7-Relatedworks}
 
@@
 %% End of file `sample-sigchi.tex'.
+
+% Prout

### 2021-01-27 implementation.tex
Started rebuttal process

The PAMELA implementation is available online. \footnote {https://github.com/openflexo-team/pamela}

\subsection{Exposed API at design-time}

The model-code integration we advocate requires facilities to encode metadata in the source code. This requires an annotation-enabled language. Such a language supports the attribute-oriented programming if its grammar allows adding custom declarative tags to annotate standard program elements. Java programming language is a good candidate, as it supports annotations.

The PAMELA API exposed to the developer mainly consists of : 1) a set of annotations; and 2) a set of unimplemented Java interfaces exposing required features.

The package \texttt{org.openflexo.pamela.annotations} package exposes the set of annotations which were presented in Subsection \ref{sub:DesignTime}.

The package \texttt{org.openflexo.pamela} contains the following feature-related java interfaces:
\begin{itemize}
    \item \texttt{AccessibleProxyObject} is the interface that PAMELA objects should extend in order to benefit from base features such as generic default implementation, containment management, notification, object graph comparison and diff/merge, visiting patterns, etc.
    \item \texttt{CloneableProxyObject} exposes features related to cloning.
    \item \texttt{DeletableProxyObject} exposes features related to deletion management.
    \item \texttt{SpecifiableProxyObject} exposes dynamic assertion checking features in the context of JML (contract management) use. 
\end{itemize}

% Generic design patterns API, used in the context of aspect programming is exposed in \texttt{org.openflexo.pamela.patterns} package. A plug-in architecture allows to enrich model with some specific design pattern. Some basic design patterns are released with PAMELA 1.6.x in the context of security (Authenticator, Authorization, SingleAccessPoint, Owner).

\subsection{PAMELA interpreter}

The package \texttt{org.openflexo.pamela.factory} contains the PAMELA interpreter implementation. The core of the interpreter is implemented in the class \texttt{ProxyMethodHandler}.

From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed. This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at run-time (support for Design by Contract, aka JML).

The PAMELA framework is a 100\% pure java (\textgreater 1.5), compilable by a classical java compiler and executable in a classical Java Virtual Machine.

% Rajouter un paragraphe sur le fait que PAMELA est une implementation du MOP (meta object protocol) ?

% Is it the good place for following links ?
%https://pamela.openflexo.org
%https://github.com/openflexo-team/pamela

### 2021-01-27 pamelaProcess.tex
Started rebuttal process

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.

\subsection{PAMELA Modeling/Development Process}

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer becomes also the modeler. In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.

% Je pense qu'ici il faudrait plutot parler du modèle. Le métamodèle PAMELA est celui représenté figure 3. Ou alors parler du metametamodele PAMELA

The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms. They include consistency checkings, which constrain their use and helps the programmer to avoid inconsistencies or errors. We have experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.

Our experience shows that introducing and reusing new concepts (1) reduce the size of the code, (2) reduce the risk of errors and (3) improve the code structure. The cycle of development between the model and the code can then be drastically reduced, leading to what we call \emph{continuous modeling}.

The code size is reduced because abstractions factorize recognized concepts so that the code using such concepts is replaced by the use of the abstraction at the right place. This also reduces the risk of errors since the code is now managed by the PAMELA framework with all the required checks. Finally, the code structure is improved since it matches the way the programmer conceptualizes (models) her/his code.

### 2021-01-27 related.tex
Started rebuttal process

Very few approaches build interfaces between MDE artifacts and source code at runtime. FAME \cite{kuhn2008fame} is  a library that keeps metamodels accessible and adaptable at runtime. Basically, FAME attaches meta-information to the objects of a running application permitting an uniform manipulation of both models and system objects at runtime. Similarly, in \cite{song2010applying} the authors use runtime models to maintain a causal connection between a model and a running systems for monitoring and control purposes. More recently, in \cite{boronat2019} the authors construct and maintain at runtime model-based views on the data manipulated in object-oriented code. Different to the aforementioned approaches, PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviors; and 2) maintaining a continuous cycle of metamodeling so that models and corresponding code remain synchronized without the need for code generation processes. In a different approach, UMPLE \cite{lethbridge2016merging} mixes programming and modeling by integrating UML constructs into languages such as Java. However, they use code generation for the runtime part of the system.

### 2021-01-27 2-Abstract.tex

@@
 \begin{abstract}
-This article presents PAMELA, an annotation-based Java Modelling framework.
-PAMELA provides a smooth integration between model and code, and enable Java developers to handle software development both at conceptual level and at source-code level, without code transformation and/or generation, and avoiding round-tripping issues.
+This article presents PAMELA, an annotation-based Java Modeling framework.
+PAMELA provides a smooth integration between model and code and enables Java developers to handle software development both at conceptual level and at source-code level, without code transformation and/or generation, avoiding round-trip related issues.
 

### 2021-01-27 6-Experimentations.tex

@@
-\section{PAMELA Experimentation}
+\section{PAMELA industrial use cases and experiments}
+\label{sec:validation}
 
-https://pamela.openflexo.org
-https://github.com/openflexo-team/pamela
+ The PAMELA framework has been successfully applied in a variety of complex programming and modeling scenarios and we continue to use it daily as part of our modeling toolbox.
+In the following, we describe three important uses cases in which PAMELA was a core component.
 
-% Parler des différentes expériences faites avec PAMELA
+%\todo[inline]{General comment: I would change the title of the section to something like, Pamela Use cases, Pamela success stories or something of the like.
+The idea here is to show that pamela is a tool that is used and solve problems.
+In that sense, the fact that pamela powers Open flexo should be highlighted.
+In the same sense, the importance of the patterns and security should be reduced and limited to the fact that pamela allows it (the point being, to cite the workshop paper).
+I believe giving details of how patterns are implemented  here will distract the reader.
+technical details, if needed would be better placed in the implementation section.}
 
-\subsection{Performance issues}
+%https://pamela.openflexo.org
+%https://github.com/openflexo-team/pamela
+
+\subsection{Openflexo infrastructure}
+
+Model Federation is an approach that provides the means to integrate multiple models conforming to different paradigms, and giving to each stakeholder a specific view adapted to its needs.
+Model federation approach is developed as a possible response to SIMF RFP (Semantic Information Modeling for Federation) by OMG (Object Modeling Group).
+This RFP (Request For Proposal) requests submissions for a standard addressing, ”federation of information across different representations, levels of abstraction, communities, organizations, viewpoints, and authorities”.
+Thus model federation allows the integration of heterogeneous models to develop new cross-concern viewpoints/models or to synchronize the models used for designing a system.
+
+Openflexo\cite{OpenflexoWebSite} is a software infrastructure providing support for model federation across multiple technological spaces.
+Conceptualization is addressed trough the proposition of a language called FML (Flexo Modeling Language), which is executable on the platform.
+Openflexo infrastructure introduces connectors (also called Technology Adapters) to support various technological spaces and paradigms.
+
+This open source initiative is now mature at the infrastructure level, and many projects and applications have been developed and powered using Openflexo infrastructure.
+More than 15 technology adapters have been developed, with various maturity stage regarding their industrialization (Microsoft Word, Excel and PowerPoint, EMF, OWL, Diagramming, JDBC, XML, OSLC, etc...)
+
+The total base of code for Openflexo infrastructure represents around 900.000 lines of Java code.
+Regarding backend modeling, the PAMELA framework is extensively used in Openflexo core as well as in most technology adapters, with very few specific implementations for properties.
+The observer/observable pattern is generally used for graphical user interfaces, which also rely on PAMELA framework.
+
+% Continuous integration ?
+
+\subsection{Formose project}
+
+The Formose ANR (french National Agency for Research) project (ANR-14-CE28-0009)\cite{FormoseWebSite} aimed to design a formally-grounded, model-based requirements engineering (RE) method for critical complex systems, supported by an open-source environment.
+The main partners were: ClearSy, LACL, Institut Mines-Telecom, OpenFlexo, and THALES.
+
+The main results of the project are a requirements modeling multi-views language, its associated design process and the development of an opensource platform called Formod\cite{FormodWebSite}, built using Openflexo infrastructure and PAMELA framework.
+The requirements modeling language is based on KAOS for goal modeling and SysML for the structural part of a system.
+The associated domain modeling language, used to describe system domain knowledge, extends the two ontology languages OWL and PLIB.
+The graphical notations are then translated into Event-B, a formal specification method supported by verification tools.
+
+The Formose method and Formod tool have been evaluated on different case studies provided by the industrial partners of the project.
+
+\subsection{SecurityPatterns experiment}
+
+A significant experiment in PAMELA is the implementation of security patterns weaved on domain code \cite{silva20}.
+In this context, the PAMELA framework is extended to include the notion of Pattern, i.e.
+a composition of multiple classes.
+Included to this experiment, the security pattern is specified by expected behavior defined and formalized by a pattern contract.
+This contract is defined by formal properties and the PAMELA framework ensures the property verification at run-time.
+
+Related to the security pattern implementation, PAMELA enables the definition of additional security behavior to existing Java code.
+Patterns are defined in PAMELA using three classes, each one representing a different conceptual level PatternFactory, PatternDefinition, PatternInstances.
+
+To declare a Pattern on existing code, pattern elements such as Pattern Stakeholders and methods need to be annotated with provided security pattern-specific annotations.
+These annotations will be discovered at run-time by the PatternFactory and stored in PatternDefinition attributes.
+
+Summarizing, implementing Patterns with PAMELA provides the ability to monitor the execution of the application code; the ability to offer extra structural and behavioral features, executed by the PAMELA interpreter; a representation of Patterns as stateful objects.
+Such objects can then evolve throughout run-time and compute assertions.
+
+%\subsection{Performance issues}
 
@@
 
-\subsection{Syntaxic issues}
+%\subsection{Syntaxic issues}
 

### 2021-01-27 architecture.tex

Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.  Basically, it is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one run-time component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{PamelaVisionV2.pdf}
    \caption{PAMELA approach for modelling}
    \label{fig:PamelaVision}
\end{figure}

\begin{center}
\begin{figure}
    %\centering
    %\includegraphics[width=1.0 \columnwidth]{PamelaMetaModel.pdf}
    \begin{tikzpicture}[scale=0.9]
    \useasboundingbox (-9,-5) rectangle (3,0);% ajustement à la main :-(
    \umlclass[x=-2,y=0]{PAMELAModel}{}{}
    \umlclass[x=-6,y=-3]{ModelEntity}{-- name \\ -- implementationInterface \\ -- implementationClass}{}
    \umlclass[x=2,y=-3]{ModelProperty}{-- name \\ -- type \\ -- cardinality \\ -- defaultValue \\ -- embeddingType \\ -- cloningStrategy \\ -- implementation }{}
    \umluniassoc[geometry=-|, arg1=baseEntities, mult1=1, pos1=0.45,  arg2=*, pos2=1.9]{PAMELAModel}{ModelEntity}
    \umluniassoc[arg1=entity, pos1=0.15, mult1=1, anchors=20 and 154, arg2=properties, mult2=*, pos2=0.75]{ModelEntity}{ModelProperty}
    \umluniassoc[anchors=154 and 20]{ModelProperty}{ModelEntity}
    \umluniassoc[mult1=entityType, pos1=0.5, arg2=0..1, pos2=0.9]{ModelProperty}{ModelEntity}
    \umluniassoc [arg=parentEntities , mult=*, pos=0.9, align= right, angle1=170, angle2=110, loopsize=2cm]{ModelEntity}{ModelEntity}
    \umluniassoc [arg=inverseProperty , mult=0..1, pos=0.9, align= left, angle1=20, angle2=80, loopsize=2cm]{ModelProperty}{ModelProperty}
    \end{tikzpicture}
    \caption{PAMELA metamodel}
    \label{fig:PamelaMetaModel}
\end{figure}
\end{center}

\subsection{Design time}
\label{sub:DesignTime}

The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of of \emph{simple} UML-like models directly on the code.

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{ModelEntity} reflects a concept and is encoded in a java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. Note that reification of \texttt{ModelEntity} is performed in a java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which will be responsible for providing custom code implementing domain or application specific behaviour. A partially implemented \texttt{abstract} java \texttt{class} may be defined as partial base implementation (conforms to implemented interface).

A \texttt{ModelEntity} also defines some properties, encoded as \texttt{ModelProperty}. A \texttt{ModelProperty} is identified by a name, a cardinality (simple or multiple) and a type, which can be a reference to another \texttt{ModelEntity}, or a Java type (a primitive or an arbitrary complex Java type). Depending on its cardinality, a \texttt{ModelProperty} is bound to a set of methods reflecting use of property.
\begin{itemize}
    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a java method defined in java interface taking no argument and returning desired value).
    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a java method taking value to be set as unique argument)
    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a java method taking value to be added as unique argument), a \emph{remover} (a java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
\end{itemize}

To summarize, in the following we list the most common java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, enventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique reindexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances

\end{itemize}

 \subsection{Runtime}
 \label{sub:RunTime}

 The aforementioned models are executed at runtime as a combination of two components: 1) plain java byte-code, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).

The main idea for the approach is to override java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed.

The PAMELA interpreter will intercept any method call for all instances of \emph{ModelEntity} and conditionally branches code execution.
 \begin{itemize}
     \item If the accessed method is part of a \emph{ModelProperty} (a getter, or a setter, etc..), and no custom implementation is defined neither in the class declared as implementation, nor in a class declared as partial implementation in the context of traits, then, execution is delegated to the related property implementation (generic code provided by the PAMELA interpreter).
     \item If the accessed method is defined in a class declared as implementation, or in a class declared as partial implementation, then this method is executed. The PAMELA API through the \mytexttt{AccessibleProxyObject} interface also provides access to generic behavior (super implementation), allowing the developer to define an overriding composition.
 \end{itemize}

This general scheme provides also an extension point allowing to instrument the code. This extension point is used in order to integrate other features such as notification management, undo/redo stack management, assertion checking at run-time (support for \emph{Design by Contract}, aka JML), and dynamic code weaving in the context of \emph{Aspect Programming}.

PAMELA model at runtime is computed dynamically, working on the classpath of launched java application, and starting from a simple java interface (or a collection of java interfaces) which is/are PAMELA-annotated. From a mathematical point of view, internal representation of the underlying model is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}. \mytexttt{@Imports} and \mytexttt{@Import} annotations allows to include some other \emph{ModelEntities} in the model. On the contrary, an annotation attribute \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that context, PAMELA model computation is a graph closure computation, starting from a collection of vertices\footnote{Model closure computation on-the-fly provides an interesting approach to deal with model fragmentation.}. A PAMELA model at runtime is represented by a \mytexttt{ModelContext}.

PAMELA instances (instances of \emph{ModelEntity}) are handled though the use of \mytexttt{ModelFactory}, which is instantiated from a \mytexttt{ModelContext}.

% Folloging section/itemize may be removed when needed 
%This composition offers many benefits: 
%\begin{itemize}
%    \item Strong coupling between model and code
%    \item Strong typing is kept, and required checks are performed by the java compiler
%    \item PAMELA framework provides interpretation of model@runtime
%    \item No need to generate POJO (plain old java objects), as their execution follow the standard semantics (less code, less bugs)
%    \item Custom implementation are provided if needed, using classical java extension points
%    \item It offers a way to intercept method calls and instrument the code
%    \item Assertions checking at runtime
%    \item Dynamic code weaving at runtime (aspect programming without compilation)
%\end{itemize}

\subsection{Additional Features}

PAMELA supports a number of advanced programming features (some already mentioned) such as:  multiple inheritance and traits, containment management, cloning, fine-grained notifications management, object graph comparison and diff/merge, visiting patterns, clipboard management, validation, support for \emph{design by contract} by integrating the Java Modeling Language (JML), Metaprogramming, Aspect Oriented Programming.

%A strong interest of the approach is that the model is encoded in java, and must be compiled. It forces the java compiler to perform required checks %for a PAMELA model encoded in a strong typed program. Execution semantics of model is fully compatible with Java semantics. Many validation rules %are automatically performed through classical java compilation, independently of underlying PAMELA execution semantics.

\subsection{Example}

Listing \ref{lst:model} shows a very basic model with two entities: \emph{Book} and \emph{Library}. Entity \emph{Book} defines two read-write single properties \emph{title} and \emph{ISBN} with single cardinality and with \texttt{String} type. Entity \emph{Book} also defines a constructor with initial \emph{title} value. Entity \emph{Library} defines a read-write multiple properties \emph{books} referencing \emph{Book} instances. Note that this code is sufficient to execute the model, while no additional line of code is required (only java interfaces and API methods are declared here).

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize, caption=Model creation, label=lst:model]
@ModelEntity
public interface Book extends AccessibleProxyObject {

  @Initializer
  public Book init(@Parameter("title")String aTitle);

  @Getter("title")
  public String getTitle();

  @Setter("title")
  public void setTitle(String aTitle);

  @Getter("ISBN")
  public String getISBN();

  @Setter("ISBN")
  public void setISBN(String value);
}

@ModelEntity
public interface Library extends AccessibleProxyObject {

  @Getter(value = "books", cardinality = Cardinality.LIST)
  public List<Book> getBooks();

  @Adder("books")
  public void addToBooks(Book aBook);

  @Remover("books")
  public void removeFromBooks(Book aBook);

  @Reindexer("books")
  public void moveBookToIndex(Book aBook, int index);

  @Finder(collection = "books", attribute = "title")
  public Book getBook(String title);
}
\end{lstlisting}
 %   \caption{A basic PAMELA model with two entities \texttt{Library} and \texttt{Book}}
 %   \label{fig:ABasicPamelaModel}
%\end{figure}

The execution and the management of this model may be performed using the following simple lines of code:

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize, caption=model execution/manipulation, label=lst:execution]
// Instantiate the meta-model
// by computing the closure of concepts graph
ModelContext modelContext 
    = ModelContextLibrary.getModelContext(Library.class);
// Instantiate the factory
ModelFactory factory  = new ModelFactory(modelContext);
// Instantiate a Library
Library myLibrary = factory.newInstance(Library.class);
myLibrary.setName("My library");
// Instantiate some Books
Book myFirstBook 
    = factory.newInstance(Book.class, "Lord of the ring");
Book anOtherBook = factory.newInstance(Book.class, "Holy bible");
myLibrary.addToBooks(myFirstBook);
myLibrary.addToBooks(anOtherBook);
\end{lstlisting}
 %   \caption{Executing a PAMELA model}
 %   \label{fig:ExecutingPamelaModel}
%\end{figure}

The first line of code instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships. This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled byte-code. A factory \texttt{ModelFactory} is then instantiated using that \texttt{ModelContext}, allowing to create \emph{Library} and \emph{Book} instances.

Custom code can be easily added to this model as we show in Listing \ref{lst:custom}. It shows how to integrate custom code to the fully interpreted \emph{Book} entity described above. The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity. Custom implementations are defined using classical java implementation/overrides scheme. Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, can not be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property. Note that this implementation references a default interpretated implementation (call to \texttt{performSuperGetter(String)} method).

%\begin{figure}
%    \centering
\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize,caption=Custom Code, label=lst:custom]
@ModelEntity
@ImplementationClass(BookImpl.class)
public interface Book extends AccessibleProxyObject {

  static final String TITLE = "title";

  @Getter(TITLE)
  String getTitle();
  // ... title property declarations ...

  void read();
}

// Provides a partial implementation for Book
public static abstract class BookImpl implements Book {

  @Override
  public String getTitle() {
    String title = performSuperGetter(TITLE);
    if (title == null) {
      return "This book has no title";
    }
    return title;
  }

  @Override
    public void read() {
      // do the job
    }
}
\end{lstlisting}
 %   \caption{Executing a PAMELA model}
 %   \label{fig:ExecutingPamelaModel}
%\end{figure}

### 2021-01-27 introduction.tex

Model-Driven Software Development focuses on managing abstract models representing the conceptual level of the solution space. These models are generally represented as various artifacts, corresponding to different languages and using different representations. But models are rarely executable and/or lack the required level of expressiveness or performance to be exploited directly in production. Thus, the solution they offer is often implemented by using code generators that produce a representation of the solution in a target programming language or framework\cite{stahl2006model}. The underlying semantics of the code to be executed is generally encoded in those code generators and can be inlined or implicitly defined by the code generation process, or may be sometimes explicit to the code generation. Figure \ref{fig:ClassicalVision} shows the classical vision for Model-Driven Engineering.

This approach, model first, then generation, raises two major issues. First, it creates an important gap between the conceptual level (the model) and the source code, where semantics may be totally hidden or implicit. Second, the synchronization of the pair models-code in a co-evolution scenario where model and source code may evolve independently (and often by different actors, e.g., architect and developer) becomes hard to maintain.

Round-trip mechanisms are commonly used to overcome those difficulties, but they are difficult to use and to maintain. Indeed, the model/code co-evolution problem remains an open subject in the software engineering research community and, consequently, software development projects often abandon the idea of maintaining the synchronization between model and source code during development process. In that context, the model is developed at the early stages of development process, used mainly to prototype software application. Then, it may be manually reviewed back at the end of development process, for documentation purposes.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{ClassicalVision.pdf}
    \caption{Classical vision for Model-Driven Engineering}
    \label{fig:ClassicalVision}
\end{figure}

Conversely, we propose a shift in the modeling paradigm in which models and code are developed together and at the same time in what we call a continuous modeling process. The PAMELA framework supports this paradigm shift by providing the means to: 1) annotate Java code with model-based annotations; and 2) interpret these annotations at runtime.

%\todo[inline]{In some sense we are loosing some of the advantages of modeling, aren't we? But let's see what reviewers say.}

% On peut éventuellement supprimer ce paragraphe
The rest of the paper is organized as follows. Section \ref{sec:pamproc} describes our approach and its associated development process. Section \ref{sec:architecture} presents the main building blocks of the PAMELA framework together with an illustrative example. Implementation details are discussed in Section \ref{sec:implementation}, followed by a description of industrial experimentation and validation cases in Section \ref{sec:validation}. We end the paper in Section \ref{sec:related} by discussing related work.

### 2021-02-02 architecture.tex
work on architecture and process (AB)

@@
-Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.
-Basically, it is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one run-time component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.
+PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one run-time component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.
 
-\begin{figure}
-    \centering
-    \includegraphics[width=1.0 \columnwidth]{PamelaVisionV2.pdf}
-    \caption{PAMELA approach for modelling}
-    \label{fig:PamelaVision}
-\end{figure}
+\subsection{Installation}
+\label{sub:installation}
+
+\todo{Maybe a word on how to install.}
 
@@
 
-The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of of \emph{simple} UML-like models directly on the code.
+The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
 
 A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}.
-A \texttt{ModelEntity} reflects a concept and is encoded in a java \texttt{interface}.
+A \texttt{ModelEntity} reflects a concept and is encoded in a Java \texttt{interface}.
 PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities.
-Note that reification of \texttt{ModelEntity} is performed in a java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods.
+Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods.
 Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which will be responsible for providing custom code implementing domain or application specific behaviour.
-A partially implemented \texttt{abstract} java \texttt{class} may be defined as partial base implementation (conforms to implemented interface).
+A partially implemented \texttt{abstract} Java \texttt{class} may be defined as partial base implementation (conforms to implemented interface).
 
@@
 \begin{itemize}
-    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a java method defined in java interface taking no argument and returning desired value).
-    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a java method taking value to be set as unique argument)
-    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a java method taking value to be added as unique argument), a \emph{remover} (a java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
+    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning desired value).
+    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking value to be set as unique argument).
+    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking value to be added as unique argument), a \emph{remover} (a Java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
 \end{itemize}
 
-To summarize, in the following we list the most common java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:
+To summarize, in the following we list the most common Java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:
 
@@
     \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declare an abstract entity.
-    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract java \texttt{class} to be used as base implementation.
+    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.
     \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
     \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value.
-May also declares cardinality, enventual inverse property, default value and some other features.
+May also declares cardinality, eventual inverse property, default value and some other features.
     \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
@@
     \item \texttt{@Reindexer(String)}: tag annotating method as unique reindexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
-    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}
-   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}
-    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute
-    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}
-    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics
-    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model
-    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances
+    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
+   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}.
+    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute.
+    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
+    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics.
+    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
+    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances.
 
@@
 
- The aforementioned models are executed at runtime as a combination of two components: 1) plain java byte-code, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).
+ The aforementioned models are executed at runtime as a combination of two components: 1) plain Java byte-code, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).
 
-The main idea for the approach is to override java dynamic binding.
+The main idea for the approach is to override Java dynamic binding.
 Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed.
@@
 
-PAMELA model at runtime is computed dynamically, working on the classpath of launched java application, and starting from a simple java interface (or a collection of java interfaces) which is/are PAMELA-annotated.
-From a mathematical point of view, internal representation of the underlying model is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}.
+PAMELA model at runtime is computed dynamically, working on the classpath of launched Java application, and starting from a simple Java interface (or a collection of Java interfaces) which is/are PAMELA-annotated.
+From a mathematical point of view, internal representation of the underlying model is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated Java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}.
 \mytexttt{@Imports} and \mytexttt{@Import} annotations allows to include some other \emph{ModelEntities} in the model.
 On the contrary, an annotation attribute \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link.
-In that context, PAMELA model computation is a graph closure computation, starting from a collection of vertices\footnote{Model closure computation on-the-fly provides an interesting approach to deal with model fragmentation.}.
+In that context, PAMELA model computation is a graph closure computation, starting from a collection of vertices.
+%\footnote{Model closure computation on-the-fly provides an interesting approach to deal with model fragmentation.} 
 A PAMELA model at runtime is represented by a \mytexttt{ModelContext}.
@@
 %    \item Strong coupling between model and code
-%    \item Strong typing is kept, and required checks are performed by the java compiler
+%    \item Strong typing is kept, and required checks are performed by the Java compiler
 %    \item PAMELA framework provides interpretation of model@runtime
-%    \item No need to generate POJO (plain old java objects), as their execution follow the standard semantics (less code, less bugs)
-%    \item Custom implementation are provided if needed, using classical java extension points
+%    \item No need to generate POJO (plain old Java objects), as their execution follow the standard semantics (less code, less bugs)
+%    \item Custom implementation are provided if needed, using classical Java extension points
 %    \item It offers a way to intercept method calls and instrument the code
@@
 
-PAMELA supports a number of advanced programming features (some already mentioned) such as:  multiple inheritance and traits, containment management, cloning, fine-grained notifications management, object graph comparison and diff/merge, visiting patterns, clipboard management, validation, support for \emph{design by contract} by integrating the Java Modeling Language (JML), Metaprogramming, Aspect Oriented Programming.
+\ab{Programmers may already reuse} a number of advanced programming features (some already mentioned) such as:  multiple inheritance and traits, containment management, cloning, fine-grained notifications management, object graph comparison and diff/merge, visiting patterns, clipboard management, validation, support for \emph{design by contract} by integrating assertions from Java Modeling Language (JML), Metaprogramming, Aspect Oriented Programming.
+\ab{Experimented programmers may extend PAMELA defining new annotations.}
 
-%A strong interest of the approach is that the model is encoded in java, and must be compiled.
-It forces the java compiler to perform required checks %for a PAMELA model encoded in a strong typed program.
+%A strong interest of the approach is that the model is encoded in Java, and must be compiled.
+It forces the Java compiler to perform required checks %for a PAMELA model encoded in a strong typed program.
 Execution semantics of model is fully compatible with Java semantics.
-Many validation rules %are automatically performed through classical java compilation, independently of underlying PAMELA execution semantics.
+Many validation rules %are automatically performed through classical Java compilation, independently of underlying PAMELA execution semantics.
 
@@
 Entity \emph{Library} defines a read-write multiple properties \emph{books} referencing \emph{Book} instances.
-Note that this code is sufficient to execute the model, while no additional line of code is required (only java interfaces and API methods are declared here).
+Note that this code is sufficient to execute the model, while no additional line of code is required (only Java interfaces and API methods are declared here).
 
@@
 
-The first line of code instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships.
+The lines 3--4 instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships.
 This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled byte-code.
@@
 The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity.
-Custom implementations are defined using classical java implementation/overrides scheme.
+Custom implementations are defined using classical Java implementation/overrides scheme.
 Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, can not be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property.

### 2021-02-02 implementation.tex
work on architecture and process (AB)

@@
-The PAMELA implementation is available online. \footnote {https://github.com/openflexo-team/pamela}
+The PAMELA implementation is available online\footnote {\url{https://github.com/openflexo-team/pamela}}.
 
@@
 
-The package \texttt{org.openflexo.pamela} contains the following feature-related java interfaces:
+The package \texttt{org.openflexo.pamela} contains the following feature-related Java interfaces:
 \begin{itemize}
@@
 
-From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the java dynamic binding.
+From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding.
 Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed.
@@
 
-The PAMELA framework is a 100\% pure java (\textgreater 1.5), compilable by a classical java compiler and executable in a classical Java Virtual Machine.
+The PAMELA framework is a 100\% pure Java (\textgreater 1.5), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.
+
+\subsection{PAMELA code base figures}
+
+\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
 

### 2021-02-02 pamelaProcess.tex
work on architecture and process (AB)

@@
 The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
+Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.
 
-\subsection{PAMELA Modeling/Development Process}
+\begin{figure}
+    \centering
+    \includegraphics[width=1.0 \columnwidth]{PamelaVisionV2.pdf}
+    \caption{PAMELA approach for modelling}
+    \label{fig:PamelaVision}
+\end{figure}
+
+%\subsection{PAMELA Modeling/Development Process}
 
@@
 The evolution rhythm between models and code is low.
-This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer becomes also the modeler.
+This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler.
 In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.
@@
 They include consistency checkings, which constrain their use and helps the programmer to avoid inconsistencies or errors.
-We have experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.
+We have first experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.
 
@@
 Finally, the code structure is improved since it matches the way the programmer conceptualizes (models) her/his code.
+
+\ab{In this presentation, we focus on the programming use of PAMELA, when programmers reuse existing annotations.
+The full power of PAMELA arises when programmers become modelers defining their own abstractions/annotations.}

### 2021-02-02 1-Main.tex
LOW / Test
work on architecture and process (AB)
(bugfix) remove a link in order to be able to compile

@@
 
-%\usepackage[colorinlisttodo]{todonotes}
+\usepackage{todonotes}
 
@@
     showtabs=false,                  
-    tabsize=2https://www.overleaf.com/project/5efee0e8ad6bc10001d47854
+    tabsize=2
 }
@@
 \DeclareTextFontCommand{\mytexttt}{\ttfamily\hyphenchar\font=45\relax}
+
+% DRAFT
+\newcommand{\ab}[1]{{\color{blue} #1}} % draft mode
 
@@
 
-\section{Approach}
+\section{Approach and Development Process}
 \label{sec:pamproc}
@@
 
-% Prout
+% Test

### 2021-02-02 architecture.tex
typos

@@
 
-To summarize, in the following we list the most common Java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:
+In the following we list the most common Java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:
 
@@
 
-PAMELA model at runtime is computed dynamically, working on the classpath of launched Java application, and starting from a simple Java interface (or a collection of Java interfaces) which is/are PAMELA-annotated.
-From a mathematical point of view, internal representation of the underlying model is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated Java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}.
-\mytexttt{@Imports} and \mytexttt{@Import} annotations allows to include some other \emph{ModelEntities} in the model.
-On the contrary, an annotation attribute \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link.
-In that context, PAMELA model computation is a graph closure computation, starting from a collection of vertices.
+PAMELA model at runtime is computed dynamically, working on the classpath of
+launched Java application, and starting from a simple Java interface (or a
+collection of Java interfaces) which is/are PAMELA-annotated. From a
+mathematical point of view, internal representation of the underlying model is
+a graph whose vertices are PAMELA \emph{ModelEntities} (annotated Java
+interface), and edges are either inheritance links or reference links (a
+property whose type is another \emph{ModelEntity}). \mytexttt{@Imports} and
+  \mytexttt{@Import} annotations allows to include some other
+  \emph{ModelEntities} in the model. On the contrary, an annotation attribute
+  \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that
+  context, PAMELA model computation is a graph closure computation, starting
+  from a collection of vertices.
+
 %\footnote{Model closure computation on-the-fly provides an interesting approach to deal with model fragmentation.} 
@@
 %    \item PAMELA framework provides interpretation of model@runtime
-%    \item No need to generate POJO (plain old Java objects), as their execution follow the standard semantics (less code, less bugs)
+%    \item No need to generate POJO (plain old Java objects), as their
+%    execution follow the standard semantics (less code, fewer bugs)
 %    \item Custom implementation are provided if needed, using classical Java extension points

### 2021-02-02 related.tex
typos

@@
-Very few approaches build interfaces between MDE artifacts and source code at runtime.
-FAME \cite{kuhn2008fame} is  a library that keeps metamodels accessible and adaptable at runtime.
-Basically, FAME attaches meta-information to the objects of a running application permitting an uniform manipulation of both models and system objects at runtime.
-Similarly, in \cite{song2010applying} the authors use runtime models to maintain a causal connection between a model and a running systems for monitoring and control purposes.
-More recently, in \cite{boronat2019} the authors construct and maintain at runtime model-based views on the data manipulated in object-oriented code.
-Different to the aforementioned approaches, PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviors; and 2) maintaining a continuous cycle of metamodeling so that models and corresponding code remain synchronized without the need for code generation processes.
-In a different approach, UMPLE \cite{lethbridge2016merging} mixes programming and modeling by integrating UML constructs into languages such as Java.
-However, they use code generation for the runtime part of the system.
+Very few approaches build interfaces between MDE artifacts and source code at
+runtime. FAME \cite{kuhn2008fame} is  a library that keeps metamodels
+accessible and adaptable at runtime. Basically, FAME attaches meta-information
+to the objects of a running application permitting a uniform manipulation of
+both models and system objects at runtime. Similarly, in
+\cite{song2010applying} the authors use runtime models to maintain a causal
+connection between a model and a running systems for monitoring and control
+purposes. More recently, in \cite{boronat2019} the authors construct and
+maintain at runtime model-based views on the data manipulated in
+object-oriented code. Different to the aforementioned approaches, PAMELA does
+not focus in reflecting and manipulating running systems but in: 1) extending
+their behaviors; and 2) maintaining a continuous cycle of metamodeling so that
+models and corresponding code remain synchronized without the need for code
+generation processes. In a different approach, UMPLE
+\cite{lethbridge2016merging} mixes programming and modeling by integrating UML
+constructs into languages such as Java. However, they use code generation for
+the runtime part of the system.

### 2021-02-04 architecture.tex
remove reference to PAMELA semantics
add a reference to other feature

@@
     \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
-    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics.
+    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as
+      embedded. %according to PAMELA semantics.
     \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
@@
 
-\ab{Programmers may already reuse} a number of advanced programming features (some already mentioned) such as:  multiple inheritance and traits, containment management, cloning, fine-grained notifications management, object graph comparison and diff/merge, visiting patterns, clipboard management, validation, support for \emph{design by contract} by integrating assertions from Java Modeling Language (JML), Metaprogramming, Aspect Oriented Programming.
+\ab{Programmers may already reuse} a number of advanced programming features
+(some already mentioned) such as: multiple inheritance and traits, containment
+management, cloning, fine-grained notifications management, object graph
+comparison and diff/merge, visiting patterns, clipboard management, validation,
+support for \emph{design by contract} by integrating assertions from Java
+Modeling Language (JML), Metaprogramming, Aspect Oriented Programming. Each of
+those features is described on a dedicated page that is reachable from the
+official web site\footnote{\url{https://pamela.openflexo.org/SNAPSHOT/}}.
 \ab{Experimented programmers may extend PAMELA defining new annotations.}

### 2021-02-04 related.tex
Updates from Overleaf

@@
-Very few approaches build interfaces between MDE artifacts and source code at
-runtime. FAME \cite{kuhn2008fame} is  a library that keeps metamodels
+The PAMELA framework can be seen as a CASE tool that focuses in the design and verification and validation phases of the software development life-cycle.
+In that sense, it present similarities to other existing object oriented CASE tools and/or language workbenches such as Kermeta~\cite{kermeta} and the Platypus~\cite{platypus} meta-case tool as they permit the definition and manipulation of models, metamodels and constraints/behavior.
+Two main features separate PAMELA from those approaches though: 1) The main artefact for the PAMELA framework is standard java code; and 2) PAMELA does not rely on code generation but on the interpretation/construction of models at runtime guided by code annotations.
+This allows for the seamlessly blending of the metamodeling/programming phases.
+
+When the observed artifact is code (e.g., for runtime analysis or verification of properties) we could see our approach as similar to some contributions appertaining to the models@runtime topic~\cite{bencomo2019models}.
+Indeed models@runtime approaches often rely  on the use of the the reflection architectural pattern~\cite{buschmann2008pattern} in order to separate core application logic from a metalevel that contains information about properties.
+This is also the case of PAMELA.
+However the focus of most models@runtime approaches is not on easing the development process but on enabling the analysis and/or the adaptation/evolution of ever running applications.
+Examples of such approaches are AC-Contrat~\cite{accontract} focusing in the Run-time verification of properties for context-aware applications.
+Ramses~\cite{ramses} focusing on dynamic adaptations,~\cite{denker2010modeling} focusing on feature analysis and~\cite{song2010applying}, which wraps running system in standard UML-like models in order to perform analysis and management task by using off-the-shelf MDE tools and techniques.
+
+More similar to PAMELA, FAME \cite{kuhn2008fame} is  a library that keeps metamodels
 accessible and adaptable at runtime. Basically, FAME attaches meta-information
 to the objects of a running application permitting a uniform manipulation of
-both models and system objects at runtime. Similarly, in
-\cite{song2010applying} the authors use runtime models to maintain a causal
-connection between a model and a running systems for monitoring and control
-purposes. More recently, in \cite{boronat2019} the authors construct and
+both models and system objects at runtime.
+
+More recently, in \cite{boronat2019} the authors construct and
 maintain at runtime model-based views on the data manipulated in

### 2021-02-08 6-Experimentations.tex
Improving related work.

@@
 The main results of the project are a requirements modeling multi-views language, its associated design process and the development of an opensource platform called Formod\cite{FormodWebSite}, built using Openflexo infrastructure and PAMELA framework.
-The requirements modeling language is based on KAOS for goal modeling and SysML for the structural part of a system.
+\todo{Say how and where PAMELA was used} The requirements modeling language is based on KAOS for goal modeling and SysML for the structural part of a system.
 The associated domain modeling language, used to describe system domain knowledge, extends the two ontology languages OWL and PLIB.
 The graphical notations are then translated into Event-B, a formal specification method supported by verification tools.
+
+\todo{Remark that it is industry ready}
 

### 2021-02-08 related.tex
Improving related work.

@@
 This is also the case of PAMELA.
-However the focus of most models@runtime approaches is not on easing the development process but on enabling the analysis and/or the adaptation/evolution of ever running applications.
+%However the focus of most models@runtime approaches is not on easing the development process but on enabling the analysis and/or the adaptation/evolution of ever running applications. 
 Examples of such approaches are AC-Contrat~\cite{accontract} focusing in the Run-time verification of properties for context-aware applications.
 Ramses~\cite{ramses} focusing on dynamic adaptations,~\cite{denker2010modeling} focusing on feature analysis and~\cite{song2010applying}, which wraps running system in standard UML-like models in order to perform analysis and management task by using off-the-shelf MDE tools and techniques.
+More similar to PAMELA, FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime.
+Basically, FAME attaches meta-information to the objects of a running application permitting a uniform manipulation of both models and system objects at runtime.
+Synchronization of changes between models and code are however limited for languages such as Java.
+In~\cite{cavarle2016dynamic} the authors present an approach to keep the (bi-directional) synchronization between feature-based models and generated code alternatives.
+They rely on Pharo\cite{pharo} reflective capabilities.
+More recently, in \cite{boronat2019} the authors construct and maintain at runtime model-based views on the data manipulated in object-oriented code.
 
-More similar to PAMELA, FAME \cite{kuhn2008fame} is  a library that keeps metamodels
-accessible and adaptable at runtime. Basically, FAME attaches meta-information
-to the objects of a running application permitting a uniform manipulation of
-both models and system objects at runtime.
+Different to the aforementioned approaches, the focus of PAMELA is not on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in~\cite{silva20}) but on: 1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel may be built incrementally without the need for code generation and thus, avoiding round-tripping issues; and 2) providing default implementations for frequently used concepts in order to ease development.
+In that sense PAMELA can be seen as a mix between classical CASE tools and the aforementioned models@runtime approaches.
 
-More recently, in \cite{boronat2019} the authors construct and
-maintain at runtime model-based views on the data manipulated in
-object-oriented code. Different to the aforementioned approaches, PAMELA does
-not focus in reflecting and manipulating running systems but in: 1) extending
-their behaviors; and 2) maintaining a continuous cycle of metamodeling so that
-models and corresponding code remain synchronized without the need for code
-generation processes. In a different approach, UMPLE
-\cite{lethbridge2016merging} mixes programming and modeling by integrating UML
+In a different approach, UMPLE\cite{lethbridge2016merging} mixes programming and modeling by integrating UML
 constructs into languages such as Java. However, they use code generation for

### 2021-02-09 6-Experimentations.tex
add few references

--- 2021-01-27 6-Experimentations.tex
@@
 
-Model Federation is an approach that provides the means to integrate multiple models conforming to different paradigms, and giving to each stakeholder a specific view adapted to its needs.
-Model federation approach is developed as a possible response to SIMF RFP (Semantic Information Modeling for Federation) by OMG (Object Modeling Group).
-This RFP (Request For Proposal) requests submissions for a standard addressing, ”federation of information across different representations, levels of abstraction, communities, organizations, viewpoints, and authorities”.
-Thus model federation allows the integration of heterogeneous models to develop new cross-concern viewpoints/models or to synchronize the models used for designing a system.
+Model Federation~\cite{Golra2016} is an approach that provides the means to
+integrate multiple models conforming to different paradigms, and giving to each
+stakeholder a specific view adapted to its needs. Model federation approach is
+developed as a possible response to SIMF RFP (Semantic Information Modeling for
+Federation)i~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
+requests submissions for a standard addressing, ”federation of information
+across different representations, levels of abstraction, communities,
+organizations, viewpoints, and authorities”. Thus model federation allows the
+integration of heterogeneous models to develop new cross-concern
+viewpoints/models or to synchronize the models used for designing a system.
 
-Openflexo\cite{OpenflexoWebSite} is a software infrastructure providing support for model federation across multiple technological spaces.
-Conceptualization is addressed trough the proposition of a language called FML (Flexo Modeling Language), which is executable on the platform.
-Openflexo infrastructure introduces connectors (also called Technology Adapters) to support various technological spaces and paradigms.
+Openflexo\cite{OpenflexoWebSite} is a software infrastructure providing support
+for model federation across multiple technological spaces. Conceptualization is
+addressed trough the proposition of a language called FML (Flexo Modeling
+Language), which is executable on the platform. Openflexo infrastructure
+introduces connectors (also called Technology Adapters) to support various
+technological spaces and paradigms.
 
-This open source initiative is now mature at the infrastructure level, and many projects and applications have been developed and powered using Openflexo infrastructure.
-More than 15 technology adapters have been developed, with various maturity stage regarding their industrialization (Microsoft Word, Excel and PowerPoint, EMF, OWL, Diagramming, JDBC, XML, OSLC, etc...)
+This open source initiative is now mature at the infrastructure level, and many
+projects and applications have been developed and powered using Openflexo
+infrastructure. More than 15 technology adapters have been developed, with
+various maturity stage regarding their industrialization (Microsoft Word, Excel
+and PowerPoint, EMF, OWL, Diagramming, JDBC, XML, OSLC, etc...)
 
-The total base of code for Openflexo infrastructure represents around 900.000 lines of Java code.
-Regarding backend modeling, the PAMELA framework is extensively used in Openflexo core as well as in most technology adapters, with very few specific implementations for properties.
-The observer/observable pattern is generally used for graphical user interfaces, which also rely on PAMELA framework.
+The total base of code for Openflexo infrastructure represents around 900.000
+lines of Java code. Regarding backend modeling, the PAMELA framework is
+extensively used in Openflexo core as well as in most technology adapters, with
+very few specific implementations for properties. The observer/observable
+pattern is generally used for graphical user interfaces, which also rely on
+PAMELA framework.
 
@@
 
-The Formose ANR (french National Agency for Research) project (ANR-14-CE28-0009)\cite{FormoseWebSite} aimed to design a formally-grounded, model-based requirements engineering (RE) method for critical complex systems, supported by an open-source environment.
-The main partners were: ClearSy, LACL, Institut Mines-Telecom, OpenFlexo, and THALES.
+The Formose ANR (french National Agency for Research) project
+(ANR-14-CE28-0009)\cite{FormoseWebSite} aimed to design a formally-grounded,
+model-based requirements engineering (RE) method for critical complex systems,
+supported by an open-source environment. The main partners were: ClearSy, LACL,
+Institut Mines-Telecom, OpenFlexo, and THALES.
 
-The main results of the project are a requirements modeling multi-views language, its associated design process and the development of an opensource platform called Formod\cite{FormodWebSite}, built using Openflexo infrastructure and PAMELA framework.
-The requirements modeling language is based on KAOS for goal modeling and SysML for the structural part of a system.
-The associated domain modeling language, used to describe system domain knowledge, extends the two ontology languages OWL and PLIB.
-The graphical notations are then translated into Event-B, a formal specification method supported by verification tools.
+The main results of the project are a requirements modeling multi-views
+language, its associated design process and the development of an opensource
+platform called Formod\cite{FormodWebSite}, built using Openflexo
+infrastructure and PAMELA framework. \todo{Say how and where PAMELA was used}
+The requirements modeling language is based on KAOS~\cite{kaos} for goal
+modeling and SysML for the structural part of a system. The associated domain
+modeling language, used to describe system domain knowledge, extends the two
+ontology languages OWL~\cite{owl} and PLIB~\cite{plib}. The graphical notations
+are then translated into Event-B~\cite{eventb}, a formal specification method supported by verification tools.
+
+\todo{Remark that it is industry ready}
 

### 2021-02-09 architecture.tex
integration of Antoine's changes (blue stuff now black)

@@
 
-\ab{Programmers may already reuse} a number of advanced programming features
+Programmers may already reuse a number of advanced programming features
 (some already mentioned) such as: multiple inheritance and traits, containment
@@
 official web site\footnote{\url{https://pamela.openflexo.org/SNAPSHOT/}}.
-\ab{Experimented programmers may extend PAMELA defining new annotations.}
+Experimented programmers may extend PAMELA defining new annotations.
 

### 2021-02-09 pamelaProcess.tex
integration of Antoine's changes (blue stuff now black)

@@
 
-\ab{In this presentation, we focus on the programming use of PAMELA, when programmers reuse existing annotations.
-The full power of PAMELA arises when programmers become modelers defining their own abstractions/annotations.}
+In this presentation, we focus on the programming use of PAMELA, when programmers reuse existing annotations.
+The full power of PAMELA arises when programmers become modelers defining their own abstractions/annotations.

### 2021-02-09 related.tex

@@
 When the observed artifact is code (e.g., for runtime analysis or verification of properties) we could see our approach as similar to some contributions appertaining to the models@runtime topic~\cite{bencomo2019models}.
-Indeed models@runtime approaches often rely  on the use of the the reflection architectural pattern~\cite{buschmann2008pattern} in order to separate core application logic from a metalevel that contains information about properties.
+Indeed models@runtime approaches often rely  on the use of the reflection architectural pattern~\cite{buschmann2008pattern} in order to separate core application logic from a metalevel that contains information about properties, types, etc.
 This is also the case of PAMELA.
 %However the focus of most models@runtime approaches is not on easing the development process but on enabling the analysis and/or the adaptation/evolution of ever running applications. 
-Examples of such approaches are AC-Contrat~\cite{accontract} focusing in the Run-time verification of properties for context-aware applications.
-Ramses~\cite{ramses} focusing on dynamic adaptations,~\cite{denker2010modeling} focusing on feature analysis and~\cite{song2010applying}, which wraps running system in standard UML-like models in order to perform analysis and management task by using off-the-shelf MDE tools and techniques.
-More similar to PAMELA, FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime.
-Basically, FAME attaches meta-information to the objects of a running application permitting a uniform manipulation of both models and system objects at runtime.
-Synchronization of changes between models and code are however limited for languages such as Java.
-In~\cite{cavarle2016dynamic} the authors present an approach to keep the (bi-directional) synchronization between feature-based models and generated code alternatives.
+Examples of such approaches are: AC-Contrat~\cite{accontract} which provides run-time verification of properties for context-aware applications; Ramses~\cite{ramses} focused on dynamic adaptations; and \cite{denker2010modeling} that performs feature analysis.~\cite{song2010applying} wraps running systems in standard UML-like models in order to perform analysis and management tasks by using off-the-shelf MDE tools and techniques whereas FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime (Synchronization of changes between models and code are however limited for languages such as Java).
+
+Different to the aforementioned approaches, the focus of PAMELA is not on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in~\cite{silva20}) but on: 1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel may be built incrementally without the need for code generation and thus, avoiding round-tripping issues; and 2) providing default implementations for frequently used abstractions in order to ease development.
+In that sense PAMELA can be seen as a mix between classical CASE tools and the aforementioned models@runtime approaches.
+
+More similar to PAMELA (as it focus on avoiding round-tripping issues) in~\cite{cavarle2016dynamic} the authors present an approach to keep the (bi-directional) synchronization between feature-based models and generated code alternatives.
 They rely on Pharo\cite{pharo} reflective capabilities.
 More recently, in \cite{boronat2019} the authors construct and maintain at runtime model-based views on the data manipulated in object-oriented code.
-
-Different to the aforementioned approaches, the focus of PAMELA is not on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in~\cite{silva20}) but on: 1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel may be built incrementally without the need for code generation and thus, avoiding round-tripping issues; and 2) providing default implementations for frequently used concepts in order to ease development.
-In that sense PAMELA can be seen as a mix between classical CASE tools and the aforementioned models@runtime approaches.
 

### 2021-02-09 architecture.tex
typos

@@
 Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods.
-Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which will be responsible for providing custom code implementing domain or application specific behaviour.
-A partially implemented \texttt{abstract} Java \texttt{class} may be defined as partial base implementation (conforms to implemented interface).
+Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
+will be responsible for providing custom code implementing domain or
+application specific behaviour. A partially implemented \texttt{abstract} Java
+\texttt{class} may be defined as partial base implementation (conforms to
+implemented interface), \jcb{as default methods in Java interfaces would be
+(since Java 8).}
 
@@
 property whose type is another \emph{ModelEntity}). \mytexttt{@Imports} and
-  \mytexttt{@Import} annotations allows to include some other
-  \emph{ModelEntities} in the model. On the contrary, an annotation attribute
+  \mytexttt{@Import} annotations allow to include some other
+  \emph{ModelEntities} in the model. An annotation attribute
   \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that
@@
 
-PAMELA instances (instances of \emph{ModelEntity}) are handled though the use of \mytexttt{ModelFactory}, which is instantiated from a \mytexttt{ModelContext}.
+PAMELA instances (instances of \emph{ModelEntity}) are handled through the use of \mytexttt{ModelFactory}, which is instantiated from a \mytexttt{ModelContext}.
 

### 2021-02-10 ResponseLetter.tex
Adding response letter template

% Title:    A LaTeX Template For Responses To a Referees' Reports
% Author:   Petr Zemek <s3rvac@gmail.com>
% Homepage: https://blog.petrzemek.net/2016/07/17/latex-template-for-responses-to-referees-reports/
% License:  CC BY 4.0 (https://creativecommons.org/licenses/by/4.0/)
\documentclass[10pt]{article}

% Allow Unicode input (alternatively, you can use XeLaTeX or LuaLaTeX)
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{xcolor}

\usepackage{microtype,xparse,tcolorbox}
\newenvironment{reviewer-comment}{}{}
\tcbuselibrary{skins}
\tcolorboxenvironment{reviewer-comment }{empty,
  left = 1em, top = 1ex, bottom = 1ex,
  borderline west = {2pt} {0pt} {black!20},
}
\ExplSyntaxOn
\NewDocumentEnvironment {response} { +m O{black!20} } {
  \IfValueT {#1} {
    \begin{reviewer-comment~}
      \setlength\parindent{2em}
      \noindent
      \ttfamily #1
    \end{reviewer-comment~}
  }
  \par\noindent\ignorespaces\color{blue}
} { \bigskip\par }

\NewDocumentCommand \Reviewer { m } {
  \section*{Comments~by~Reviewer~#1}
}

\NewDocumentCommand \ToolComments { m } {
  \section*{Tool~Comments~by~Reviewer~#1}
}

\ExplSyntaxOff
\AtBeginDocument{\maketitle\thispagestyle{empty}\noindent}

% You can get probably get rid of these definitions:
\newcommand\meta[1]{$\langle\hbox{#1}\rangle$}
\newcommand\PaperTitle[1]{``\textit{#1}''}

\title{Response letter to Manuscript SCICO-D-20-00176 entitled "PAMELA: an Annotation-Based Java Modeling Framework"}
%\author{Author1 \and Author2 \and Author3}
\date{\today}

\begin{document}

Dear Editor,

\bigskip
Thanks for your message of January 9th, 2021, in which you suggested preparing a major revision of our manuscript  \textbf{PAMELA: an Annotation-Based Java Modeling Framework}.

\bigskip
We are grateful to the reviewers for their interest, effort, and suggestions. We detail in the following pages how we have addressed the reviewers' concerns; we are confident that the paper is now stronger, and hope that you will agree. We have organized the responses to the comments on a per-reviewer basis, focusing, as suggested, on:

\begin{itemize}
\item Improving the readability of the manuscript. 
\item Improving tool documentation and examples.
\item Provide additional experiments \& evaluation.
\end{itemize}

Moreover, the paper has been thoroughly revised to fix minor issues and to give the additional clarifications required by the reviewers. Please let us know if you need any further information or clarification.

\bigskip
Sincerely yours,

\bigskip
Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador and Antoine

\pagebreak

\Reviewer{\#1}
\begin{response}{They then briefly discuss implementation details, in a single page; and then discuss 'industrial use cases and experiments'.

Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
}
  Response to the first comment.
\end{response}
\begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
  Response to the second comment.
\end{response}

\begin{response}{What would also be interesting is a general discussion on performance of your tool, gains by using your tool, ...
}
  Response to n-ieme comment.
\end{response}

\begin{response}{Your reference list is incomplete and inconsistent. Formod or Formosa? When were URLs accessed (add date). DOIs are missing. Ref. 5 should have the ISoLA track name and editors added. Ref. 7 is missing publisher, journal, conference. Inconsistent use of first names, first names + initials, first initial.}
  Response to n-ieme comment.
\end{response}

\begin{response}{The paper highlights include "Contract programming with assertions [remove the plural] checking at runtime", yet that is not mentioned in the paper at all.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{That also brings up the question: what about correctness of your code? Any evidence for that?}
  Response to n-ieme comment.
\end{response}

\begin{response}{p10 l32+: very little detail or quantification on the tool's code base. Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
  Response to n-ieme comment.
\end{response}

\begin{response}{p10 l32+: very little detail or quantification on the tool's code base. Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
  Response to n-ieme comment.
\end{response}

\ToolComments{\#1}

\begin{response}{gives "git clone git@github.com:openflexo-team/pamela.git", which (even when using my own username) does not work. However, "git clone https://github.com/openflexo-team/pamela.git" immediately does. Why not state the latter?
}
  Response to n-ieme comment.
\end{response}

\begin{response}{how to actually build the environment is not stated for this option.}
  Response to n-ieme comment.
\end{response}

\begin{response}{Run tests Unit tests located in src/test/java in pamela-core are a good starting point to discover the framework." No further information is given. Apparently, the user has to figure out him/herself how to run the tests. Documentation is lacking here.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{Make your first model." And then what?
}
  Response to n-ieme comment.
\end{response}

\begin{response}{Summarizing: the *getting started* documentation is incomplete and certainly not suitable for getting started. The page "https://pamela.openflexo.org/SNAPSHOT/example.html" gives a basic example, but the relation between that and the getting started page is not made explicit, nor do the two in their current shape provide a user with a proper getting started, allowing him/her to at least (1) install the framework, (2) create a basic example, (3) use the framework with that basic example. This needs to be improved before an evaluation of the software can be completed.
}
  Response to n-ieme comment.
\end{response}

\pagebreak

\Reviewer{\#2}
\begin{response}{scenarios are missing. As a developer I don't know in which cases it is relevant to use this tool.}
  Response to the first comment.
\end{response}

\begin{response}{how to use the tool is not describe properly. It is not clear if the developers need to write all the annotations. If an application is built with PAMELA, what is the link between the interfaces and the concrete classes. Indeed, at one point, code corresponding to the behaviour will be needed; and not only the accessors to navigate the structure of the model.}
  Response to n-ieme comment.
\end{response}

\begin{response}{how does such "code" evolve with time? Authors have used PAMELA for years, how such applications resist with time?}
  Response to n-ieme comment.
\end{response}

\begin{response}{section 2 is too abstract to be very useful}
  Response to n-ieme comment.
\end{response}

\begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
  Response to n-ieme comment.
\end{response}

\begin{response}{the partially implemented class concept is not clear.}
  Response to n-ieme comment.
\end{response}

\begin{response}{the additional features seem relevant, but they are not described. Either the authors describe them a bit more or they remove the section.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 3.4 it is not clear who writes what and at which step of the development process.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 5, the use of PAMELA in the two projects is not clear.}
  Response to n-ieme comment.
\end{response}

\begin{response}{references are missing. For example Kermeta done, Platypus (from Alain Plantec) done, work of Glenn Cavarle and others from the model @runtime community.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 6, the authors write: "PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviours; and 2) maintaining a continuous cycle of metamodeling... These two points are not clear previously in the paper.
}
  Response to n-ieme comment.
\end{response}

\ToolComments{\#2}

\begin{response}{The link in the paper is not clickable and copy-and-paste does not work (because of latex?) Link found: https://github.com/openflexo-team/pamela}
  Response to n-ieme comment.
\end{response}

\begin{response}{Installation
The readme of the GitHub page should link to the installation page found in https://pamela.openflexo.org. It is unclear we must visit this web site to install the tool.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{
What I have done

Create a new project in Intellij Idea with gradle
Add the line found in \url{https://pamela.openflexo.org/SNAPSHOT/getting_started.html}

- Test 1: Create a model with only one entity without error

- Test 2: Create the example of the paper

--->Bug: in Listing2, the authors execute Line9 ".setName" but the method does not exists
Else, the example works.

}
  Response to n-ieme comment.
\end{response}

\end{document}

### 2021-02-10 implementation.tex
Adding response letter template

@@
 
-\subsection{PAMELA code base figures}
+\subsection{PAMELA code base metrics}
 
-\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
+\sg{
+
+PAMELA implementation is modularized.
+
+\begin{itemize}
+    \vspace{-0.2cm}\item Core implementation (\texttt{pamela-core}) provides all base features for PAMELA.
+It contains 20k lines of code involving 184 classes.
+This code is covered with unit test (6k lines of code, 112 classes).
+Reached code coverage is about 66\%.
+    \vspace{-0.2cm}\item \texttt{pamela-security-patterns} is an add-on library containing some security-pattern implementations.
+    \vspace{-0.2cm}\item \texttt{pamela-perf-tests} contains benchmarking tools whose purpose is to quantifiate PAMELA performances. 
+\end{itemize}
+
+}
+%\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
+
+\subsection{On performance issues}
+
+\sg{
+Compared to a basic POJO (Plain Object Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.
+
+A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developped to help measure that overhead.
+We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value).
+From this simple model, two java implementations have been derived using code generation.
+The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpretated (it uses PAMELA framework and defines only interfaces and API methods).
+
+Unsurprisingly, we quantifiate a significant CPU overhead (from x20).
+This is due to the fact that \texttt{MethodHandler} mechanism is fully interpretated and contains many hooks, and cannot be compared to a fully-compiled and optimized Java dynamic binding.
+As well, memory footprint is increased from a x5 factor.
+
+These conclusions must be moderated by the fact that this overhead only applies to a very small portion of code (mostly property accessors methods).
+A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer.
+Using PAMELA won't increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case).
+As we will see in next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.
+
+Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc...).
+
+}
 

### 2021-02-10 pamelaProcess.tex
Adding response letter template

@@
 PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization.
+\sg{Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed.}.
 The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
@@
 This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations.
-The evolution rhythm between models and code is low.
-This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler.
+The evolution rythm between models and code is low.
+This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rythm when the programmer also becomes the modeler.
 In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.
@@
 
+\sg{
+Here are various conceivable scenarios for PAMELA use:
+\begin{itemize}
+    \vspace{-0.2cm}\item \textbf{Programming use} : programmers reuse existing annotations and write model and code at the same time (the modelers and programmers share the same artifacts : the Java code).
+This scenario includes the case where the model (made by the modelers) is pre-existing.
+    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing existing code by abstract method declarations in Java interface, reducing risks of errors.
+    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (eg Security Patterns) which offers code weaving at runtime, as well as runtime monitoring.
+    \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extends PAMELA with their own annotations, their own implementations, their own patterns, etc...
+\end{itemize}
+}
+
 In this presentation, we focus on the programming use of PAMELA, when programmers reuse existing annotations.

### 2021-02-11 ResponseLetter.tex
Adding comments to the response letter

@@
 \usepackage{microtype,xparse,tcolorbox}
-\newenvironment{reviewer-comment}{}{}
+\newenvironment{reviewer-comment }{}{}
 \tcbuselibrary{skins}
@@
   \IfValueT {#1} {
-    \begin{reviewer-comment~}
+    \begin{reviewer-comment~ }
       \setlength\parindent{2em}
@@
       \ttfamily #1
-    \end{reviewer-comment~}
+    \end{reviewer-comment~ }
   }
@@
 Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
-}
-  Response to the first comment.
+} Response to the first comment.
 \end{response}
 \begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
-  Response to the second comment.
+
+We fully agree with the reviewer.
+We have expanded the related work section and compared our work with existing metamodeling and models@runtime approaches.
+Moreover, we have included a number of new references as suggested by Reviewer 2.
+
 \end{response}
@@
 Inconsistent use of first names, first names + initials, first initial.}
-  Response to n-ieme comment.
+We have expanded our reference list and fixed all (detected) insconsistencies.
 \end{response}
@@
 \begin{response}{That also brings up the question: what about correctness of your code? Any evidence for that?}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{p10 l32+: very little detail or quantification on the tool's code base.
-Surely you can say more about its structure, size of components, ...
-other interesting and relevant observations on the codebase?}
-  Response to n-ieme comment.
+PAMELA have been thoroughly tested. Indeed, the test suite (which is available in package XXX) reaches a coverage of XX\% of the code.
 \end{response}
@@
 }
-  Response to n-ieme comment.
+
+We apologize for the inconvenience. We have replaced "git clone git@github.com:openflexo-team/pamela.git" for "git clone https://github.com/openflexo-team/pamela.git" as suggested.
 \end{response}

### 2021-02-11 ResponseLetter.tex
Adding answers to the response letter

@@
 \begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
-
 We fully agree with the reviewer.
@@
 Moreover, we have included a number of new references as suggested by Reviewer 2.
-
 \end{response}
@@
 \begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
-  Response to n-ieme comment.
+We have remove this reference to PAMELA semantics in the paper. Our intention was to refer to the PAMELA metamodel.
 \end{response}
@@
 \begin{response}{the partially implemented class concept is not clear.}
-  Response to n-ieme comment.
+The partially implemented class concept is very similar to the default methods introduced in Java 8 (which were not available when PAMELA was first created) permitting a sort of multiple inheritance.
+We have clarified this in the paper.
 \end{response}
@@
 
-\begin{response}{references are missing. For example Kermeta done, Platypus (from Alain Plantec) done, work of Glenn Cavarle and others from the model @runtime community.}
-  Response to n-ieme comment.
+\begin{response}{references are missing. For example Kermeta done, Platypus (from Alain Plantec), work of Glenn Cavarle and others from the model @runtime community.}
+We have extended the related work section in order to discuss the suggested related works (among others).
 \end{response}
@@
 }
-  Response to n-ieme comment.
+We have removed that part as it was indeed misleading and not discussed in that form previously in the paper. We have rephrased as follows:
+
+PAMELA is not focused on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in [20]) but on:  1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel  may  be  built  incrementally  without  the  need  for  code  generation and thus,  avoiding round-tripping issues;  and 2) providing default implementations for frequently used abstractions in order to ease development.
 \end{response}
@@
 \begin{response}{The link in the paper is not clickable and copy-and-paste does not work (because of latex?) Link found: https://github.com/openflexo-team/pamela}
-  Response to n-ieme comment.
+Thanks for pointing this out. We have modified the latex settings so that this link (and all the others throughout the paper) is now clickable.
 \end{response}

### 2021-02-12 6-Experimentations.tex
Updates from Overleaf

@@
 
+\sg{
+The full Openflexo infrastructure is composed of about 50 components.
+In most components, PAMELA framework is extensively used.
+As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagramming features) is composed of 998 classes.
+159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}.
+}
+
 The total base of code for Openflexo infrastructure represents around 900.000
@@
 PAMELA framework.
+
+\sg{
+An interesting experimentation has been done in the context of Openflexo development process.
+When Pamela has been integrated to the code base, a big portion of former legacy code has gradually and iteratively been migrated to Pamela.
+Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration.
+In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.
+}
+
+\sg{
+Pamela implementation is now really stable and mature.
+According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.
+}
 

### 2021-02-12 architecture.tex
Adding response letter template

@@
 
-\subsection{Installation}
-\label{sub:installation}
+%\subsection{Installation}
+%\label{sub:installation}
 
-\todo{Maybe a word on how to install.}
+%\todo{Maybe a word on how to install.}
+% We decided not to do it
 
@@
 those features is described on a dedicated page that is reachable from the
-official web site\footnote{\url{https://pamela.openflexo.org/SNAPSHOT/}}.
+official web site\footnote{\url{https://pamela.openflexo.org}}.
 Experimented programmers may extend PAMELA defining new annotations.

### 2021-02-12 6-Experimentations.tex
sync with overleaf

@@
 Related to the security pattern implementation, PAMELA enables the definition of additional security behavior to existing Java code.
-Patterns are defined in PAMELA using three classes, each one representing a different conceptual level PatternFactory, PatternDefinition, PatternInstances.
+Patterns are defined in PAMELA using three classes, each one representing a different conceptual level \texttt{PatternFactory}, \texttt{PatternDefinition}, \texttt{PatternInstance}.
 
 To declare a Pattern on existing code, pattern elements such as Pattern Stakeholders and methods need to be annotated with provided security pattern-specific annotations.
-These annotations will be discovered at run-time by the PatternFactory and stored in PatternDefinition attributes.
+These annotations will be discovered at run-time by the \texttt{PatternFactory} and stored in \texttt{PatternDefinition} attributes.
 

### 2021-02-12 architecture.tex
sync with overleaf

@@
 Library myLibrary = factory.newInstance(Library.class);
-myLibrary.setName("My library");
 // Instantiate some Books

### 2021-02-12 implementation.tex
sync with overleaf

@@
 
-The PAMELA framework is a 100\% pure Java (\textgreater 1.5), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.
+The PAMELA framework is a 100\% pure Java ($\geq$ /1.81.8), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.
 

### 2021-02-12 pamelaProcess.tex
sync with overleaf

@@
 PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization.
-\sg{Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed.}.
-The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
-Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.
+\sg{Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted).} The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
+Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.
 
@@
     \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing existing code by abstract method declarations in Java interface, reducing risks of errors.
-    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (eg Security Patterns) which offers code weaving at runtime, as well as runtime monitoring.
+    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (eg Security Patterns \cite{silva20}) which offers code weaving at runtime, as well as runtime monitoring.
     \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extends PAMELA with their own annotations, their own implementations, their own patterns, etc...

### 2021-02-15 ResponseLetter.tex
Adding to the response letter

@@
 Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
-} Response to the first comment.
+} We agree with the reviewer.
+We have extended the manuscript in order to provide more details about PAMELA itself, its usage and its performance.
+Concretely, we have added code base metrics in Section 4.3., a performance evaluation and discussion in Section 4.4, and more details on the role PAMELA took in the industrial use cases and experiments described in Section 5.
 \end{response}
+
 \begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
@@
 \begin{response}{What would also be interesting is a general discussion on performance of your tool, gains by using your tool, ...
-}
-Response to n-ieme comment.
+} We discuss PAMELA performance in a new Section 4.4.
+We have created a workbench consisting in an small model (4 entities containing 4-5 properties) and measured how well PAMELA fares against a basic POJO implementation when dealing with large instance models.
 \end{response}
@@
 \begin{response}{The paper highlights include "Contract programming with assertions [remove the plural] checking at runtime", yet that is not mentioned in the paper at all.
-}
-  Response to n-ieme comment.
+} Thank you for pointing this out. We have removed this point from the highlights as it is not on the focus of the paper.
 \end{response}
@@
 \begin{response}{That also brings up the question: what about correctness of your code? Any evidence for that?}
-PAMELA have been thoroughly tested. Indeed, the test suite (which is available in package XXX) reaches a coverage of XX\% of the code.
+PAMELA have been thoroughly tested.
+The core PAMELA implementation is accompained with a battery of unit tests (6k lines of code and 112 classes) reaching a coverage of 66\%.
+We have modified the paper in order to add this information in a new Section 4.3.
 \end{response}

### 2021-02-15 ResponseLetter.tex
answers, answers!

@@
 other interesting and relevant observations on the codebase?}
-  Response to n-ieme comment.
+The new Section 4.3. provides details about the PAMELA code base.
+\end{response}
+
+\begin{response}{Detailed comments}
+We have fixed all the issued highlighted in the detailed comments.
 \end{response}
@@
 \begin{response}{how to actually build the environment is not stated for this option.}
-  Response to n-ieme comment.
+We apologize for the inconvenience. The PAMELA documentation and web site have been updated in order to solve the problem.
 \end{response}
@@
 }
-  Response to n-ieme comment.
+We have refactored the PAMELA documentation on the web site so that it redirects to the getting started guide.
 \end{response}
@@
 }
-  Response to n-ieme comment.
+We apologize again for the lack of clarity in the documentation.
+It has been refactored so that the starting point is clear and the reader is guided step by step to the realization/usage of a first model.
 \end{response}
@@
 \begin{response}{scenarios are missing. As a developer I don't know in which cases it is relevant to use this tool.}
-  Response to the first comment.
+We have rewritten Section 2 so that it includes now a list and brief description of typical usage scenarios for the PAMELA framework (e.g., coding-metamodeling blending, aspect-oriented programming or legacy code re-engineering).
+We also state that the standard PAMELA user will reuse existing annotation while more advanced users may become (meta) modelers and provide their own abstractions in a continous modeling approach.
 \end{response}
@@
 Indeed, at one point, code corresponding to the behaviour will be needed; and not only the accessors to navigate the structure of the model.}
-  Response to n-ieme comment.
+We have modified Section 2 to clarify that standard PAMELA users will reuse existing annotations which are accompanied with default implementations in the PAMELA interpreter.
+Additional details are given in Sections 3 and 4 (and in the online documentation: \url{https://pamela.openflexo.org/SNAPSHOT/pamela-core/3-metaprogramming_support.html}) See Listing 3 for an example of how the default implementation can be replaced by custom ones.
 \end{response}
@@
 \begin{response}{how does such "code" evolve with time? Authors have used PAMELA for years, how such applications resist with time?}
-  Response to n-ieme comment.
+PAMELA helps factorizing code corresponding to recurrent concepts.
+In that sense its use does reduce the size and duplication of code, decreases the number of bugs and improves the code structure.
+All this makes final applications that use PAMELA (more) easily maintainable.
+We briefly discuss in Section 5.1. how integrating PAMELA in the OpenFlexo tool helped removing duplicated code and eliminating bugs in the latter.
 \end{response}

### 2021-02-15 ResponseLetter.tex
And my final pass on the response letter is done...

@@
 \usepackage{xcolor}
+\usepackage{todonotes}
 
@@
 \begin{itemize}
-\item Improving the readability of the manuscript. 
+\item Providing further details on the usage scenarios of PAMELA.
+\item Provide additional details about experiments \& evaluation.
 \item Improving tool documentation and examples.
-\item Provide additional experiments \& evaluation.
 \end{itemize}
@@
 PAMELA have been thoroughly tested.
-The core PAMELA implementation is accompained with a battery of unit tests (6k lines of code and 112 classes) reaching a coverage of 66\%.
+The core PAMELA implementation is accompanied with a battery of unit tests (6k lines of code and 112 classes) reaching a coverage of 66\%.
 We have modified the paper in order to add this information in a new Section 4.3.
@@
 }
-  Response to n-ieme comment.
+ Answer to question X \todo{I do not know what to answer here.}
 \end{response}
@@
 We apologize again for the lack of clarity in the documentation.
-It has been refactored so that the starting point is clear and the reader is guided step by step to the realization/usage of a first model.
+We have refactored it so that the starting point is clear and the reader is guided step by step to the realization/usage of a first model.
 \end{response}
@@
 We have rewritten Section 2 so that it includes now a list and brief description of typical usage scenarios for the PAMELA framework (e.g., coding-metamodeling blending, aspect-oriented programming or legacy code re-engineering).
-We also state that the standard PAMELA user will reuse existing annotation while more advanced users may become (meta) modelers and provide their own abstractions in a continous modeling approach.
+We also state that the standard PAMELA user will reuse existing annotation while more advanced users may become (meta) modelers and provide their own abstractions in a continuous modeling approach.
 \end{response}
@@
 If an application is built with PAMELA, what is the link between the interfaces and the concrete classes.
-Indeed, at one point, code corresponding to the behaviour will be needed; and not only the accessors to navigate the structure of the model.}
+Indeed, at one point, code corresponding to the behavior will be needed; and not only the accessors to navigate the structure of the model.}
 We have modified Section 2 to clarify that standard PAMELA users will reuse existing annotations which are accompanied with default implementations in the PAMELA interpreter.
@@
 \begin{response}{section 2 is too abstract to be very useful}
-  Response to n-ieme comment.
+Please, see the two first answers above where we explain how we have extended Section 2.
 \end{response}
@@
 \begin{response}{the additional features seem relevant, but they are not described. Either the authors describe them a bit more or they remove the section.}
-  Response to n-ieme comment.
+Describing the additional features would require quite a bit of text and the risk of taking the focus of the reader away from the main features of the framework.
+We have thus decided to keep the section short but alleviate the problem pointed out by the reviewer by adding a link to the PAMELA web site where all these additional features are listed and extensively described.
 \end{response}
@@
 \begin{response}{in section 3.4 it is not clear who writes what and at which step of the development process.}
-  Response to n-ieme comment.
+With respect to Listing 1 and 2, everything is written by the user which uses default PAMELA annotations so that the models it is defining (together with the code) may be created and manipulated directly (i.e., without the need for providing implementation for common tasks like model instance creation or attribute access).
+It is then the PAMELA interpreter that takes care of providing this aforementioned default behavior (in order to ease the discussion, Section 3.4.
+has been enriched with links to the Figure 2 where the PAMELA approach is depicted).
+Finally, the end of Section 3 shows how the user may write custom code that would replace the default behavior provided by the PAMELA interpreter (see Listing 3).
 \end{response}
@@
 \begin{response}{in section 5, the use of PAMELA in the two projects is not clear.}
-  Response to n-ieme comment.
+We have extended Section 5 in order to add details with respect to the usage of PAMELA.
+Notably, in Section 5.1.
+many details have been included describing how PAMELA is extensively used within the OpenFlexo tool for every aspect involving model entities, including graphical user interfaces.
+Additionally, we report on how integrating PAMELA in OpenFlexo helped reducing code duplication and eliminating bugs.
 \end{response}
@@
 
-\begin{response}{in section 6, the authors write: "PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviours; and 2) maintaining a continuous cycle of metamodeling...
+\begin{response}{in section 6, the authors write: "PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviors; and 2) maintaining a continuous cycle of metamodeling...
 These two points are not clear previously in the paper.
@@
 PAMELA is not focused on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in [20]) but on:  1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel  may  be  built  incrementally  without  the  need  for  code  generation and thus,  avoiding round-tripping issues;  and 2) providing default implementations for frequently used abstractions in order to ease development.
+\end{response}
+
+\begin{response}{Minor remarks}
+We have fixed all minor remarks.
 \end{response}
@@
 }
-  Response to n-ieme comment.
+Thanks again. We completely agree. We have included a link to the PAMELA web site in the readme of the GitHub repository.
 \end{response}
@@
 }
-  Response to n-ieme comment.
+We had indeed overlooked the bug on line 9. Thanks for pointing it out. We have removed that line from the listing.
 \end{response}

### 2021-02-15 6-Experimentations.tex
typos + homogeneity

@@
 Included to this experiment, the security pattern is specified by expected behavior defined and formalized by a pattern contract.
-This contract is defined by formal properties and the PAMELA framework ensures the property verification at run-time.
+This contract is defined by formal properties and the PAMELA framework ensures the property verification at runtime.
 
@@
 To declare a Pattern on existing code, pattern elements such as Pattern Stakeholders and methods need to be annotated with provided security pattern-specific annotations.
-These annotations will be discovered at run-time by the \texttt{PatternFactory} and stored in \texttt{PatternDefinition} attributes.
+These annotations will be discovered at runtime by the \texttt{PatternFactory} and stored in \texttt{PatternDefinition} attributes.
 
 Summarizing, implementing Patterns with PAMELA provides the ability to monitor the execution of the application code; the ability to offer extra structural and behavioral features, executed by the PAMELA interpreter; a representation of Patterns as stateful objects.
-Such objects can then evolve throughout run-time and compute assertions.
+Such objects can then evolve throughout runtime and compute assertions.
 

### 2021-02-15 architecture.tex
typos + homogeneity

@@
-PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one run-time component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.
+PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one runtime component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.
 
@@
 
- The aforementioned models are executed at runtime as a combination of two components: 1) plain Java byte-code, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).
+ The aforementioned models are executed at runtime as a combination of two components as illustrated by  the right part of Figure~\ref{fig:PamelaVision}: 1) plain Java bytecode, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).
 
@@
 This general scheme provides also an extension point allowing to instrument the code.
-This extension point is used in order to integrate other features such as notification management, undo/redo stack management, assertion checking at run-time (support for \emph{Design by Contract}, aka JML), and dynamic code weaving in the context of \emph{Aspect Programming}.
+This extension point is used in order to integrate other features such as notification management, undo/redo stack management, assertion checking at runtime (support for \emph{Design by Contract}, aka JML), and dynamic code weaving in the context of \emph{Aspect Programming}.
 
@@
 The lines 3--4 instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships.
-This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled byte-code.
+This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled bytecode.
 A factory \texttt{ModelFactory} is then instantiated using that \texttt{ModelContext}, allowing to create \emph{Library} and \emph{Book} instances.

### 2021-02-15 implementation.tex
typos + homogeneity

@@
 Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed.
-This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at run-time (support for Design by Contract, aka JML).
+This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).
 

### 2021-02-15 related.tex
typos + homogeneity

@@
 %However the focus of most models@runtime approaches is not on easing the development process but on enabling the analysis and/or the adaptation/evolution of ever running applications. 
-Examples of such approaches are: AC-Contrat~\cite{accontract} which provides run-time verification of properties for context-aware applications; Ramses~\cite{ramses} focused on dynamic adaptations; and \cite{denker2010modeling} that performs feature analysis.~\cite{song2010applying} wraps running systems in standard UML-like models in order to perform analysis and management tasks by using off-the-shelf MDE tools and techniques whereas FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime (Synchronization of changes between models and code are however limited for languages such as Java).
+Examples of such approaches are: AC-Contrat~\cite{accontract} which provides runtime verification of properties for context-aware applications; Ramses~\cite{ramses} focused on dynamic adaptations; and \cite{denker2010modeling} that performs feature analysis.~\cite{song2010applying} wraps running systems in standard UML-like models in order to perform analysis and management tasks by using off-the-shelf MDE tools and techniques whereas FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime (Synchronization of changes between models and code are however limited for languages such as Java).
 

### 2021-02-16 implementation.tex
Antoine's changes

@@
 It contains 20k lines of code involving 184 classes.
-This code is covered with unit test (6k lines of code, 112 classes).
+This code is covered with unit tests (6k lines of code, 112 classes).
 Reached code coverage is about 66\%.
     \vspace{-0.2cm}\item \texttt{pamela-security-patterns} is an add-on library containing some security-pattern implementations.
-    \vspace{-0.2cm}\item \texttt{pamela-perf-tests} contains benchmarking tools whose purpose is to quantifiate PAMELA performances. 
+    \vspace{-0.2cm}\item \texttt{pamela-perf-tests} contains benchmarking tools whose purpose is to quantify PAMELA performances. 
 \end{itemize}

### 2021-02-16 pamelaProcess.tex
Antoine's changes

@@
 This scenario includes the case where the model (made by the modelers) is pre-existing.
-    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing existing code by abstract method declarations in Java interface, reducing risks of errors.
-    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (eg Security Patterns \cite{silva20}) which offers code weaving at runtime, as well as runtime monitoring.
-    \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extends PAMELA with their own annotations, their own implementations, their own patterns, etc...
+    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing this code by abstract method declarations in Java interface, reducing risks of errors.
+    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (e.g., Security Patterns \cite{silva20}) which offers code weaving at runtime, and runtime monitoring.
+    \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extend PAMELA with their own annotations, implementations or patterns.
 \end{itemize}

### 2021-02-16 6-Experimentations.tex
typos, harmonization of vocabulary, remove colors, ...

@@
 developed as a possible response to SIMF RFP (Semantic Information Modeling for
-Federation)i~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
+Federation)~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
 requests submissions for a standard addressing, ”federation of information
@@
 
-\sg{
 The full Openflexo infrastructure is composed of about 50 components.
@@
 159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}.
-}
 
@@
 
-\sg{
 An interesting experimentation has been done in the context of Openflexo development process.
@@
 In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.
-}
 
-\sg{
 Pamela implementation is now really stable and mature.
 According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.
-}
 

### 2021-02-16 architecture.tex
typos, harmonization of vocabulary, remove colors, ...

@@
 \texttt{class} may be defined as partial base implementation (conforms to
-implemented interface), \jcb{as default methods in Java interfaces would be
-(since Java 8).}
+implemented interface), as default methods in Java interfaces would be
+(since Java 8).
 
@@
 %    \item PAMELA framework provides interpretation of model@runtime
-%    \item No need to generate POJO (plain old Java objects), as their
+%    \item No need to generate POJO (Plain Old Java Objects), as their
 %    execution follow the standard semantics (less code, fewer bugs)

### 2021-02-16 implementation.tex
typos, harmonization of vocabulary, remove colors, ...

@@
 
-The PAMELA framework is a 100\% pure Java ($\geq$ /1.81.8), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.
+The PAMELA framework is a 100\% pure Java ($\geq$1.8), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.
 
 \subsection{PAMELA code base metrics}
-
-\sg{
 
@@
 
-}
 %\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
@@
 
-\sg{
-Compared to a basic POJO (Plain Object Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.
+Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.
 
@@
 
-}
-
 % Rajouter un paragraphe sur le fait que PAMELA est une implementation du MOP (meta object protocol) ?

### 2021-02-16 pamelaProcess.tex
typos, harmonization of vocabulary, remove colors, ...

@@
 PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization.
-\sg{Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted).} The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
+Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted).
+The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
 Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.
@@
 
-\sg{
 Here are various conceivable scenarios for PAMELA use:
@@
 \end{itemize}
-}
 

### 2021-02-16 related.tex
typos, harmonization of vocabulary, remove colors, ...

@@
 The PAMELA framework can be seen as a CASE tool that focuses in the design and verification and validation phases of the software development life-cycle.
-In that sense, it present similarities to other existing object oriented CASE tools and/or language workbenches such as Kermeta~\cite{kermeta} and the Platypus~\cite{platypus} meta-case tool as they permit the definition and manipulation of models, metamodels and constraints/behavior.
+In that sense, it presents similarities to other existing object oriented CASE tools and/or language workbenches such as Kermeta~\cite{kermeta} and the Platypus~\cite{platypus} meta-case tool as they permit the definition and manipulation of models, metamodels and constraints/behavior.
 Two main features separate PAMELA from those approaches though: 1) The main artefact for the PAMELA framework is standard java code; and 2) PAMELA does not rely on code generation but on the interpretation/construction of models at runtime guided by code annotations.

### 2021-02-17 2-Abstract.tex
abstract after Antidote

@@
 \begin{abstract}
-This article presents PAMELA, an annotation-based Java Modeling framework.
-PAMELA provides a smooth integration between model and code and enables Java developers to handle software development both at conceptual level and at source-code level, without code transformation and/or generation, avoiding round-trip related issues.
+This article presents PAMELA, an annotation-based Java modeling framework.
+PAMELA provides a smooth integration between model and code and enables Java developers to handle software development both at conceptual level and at source-code level, without code transformation and/or generation, avoiding round-trip-related issues.
 

### 2021-02-17 introduction.tex
introduction after Antidote

@@
 Thus, the solution they offer is often implemented by using code generators that produce a representation of the solution in a target programming language or framework\cite{stahl2006model}.
-The underlying semantics of the code to be executed is generally encoded in those code generators and can be inlined or implicitly defined by the code generation process, or may be sometimes explicit to the code generation.
+The underlying semantics of the code to be executed is generally encoded in those code generators and can be inlined or implicitly defined by the code generation process, or may sometimes be explicit to the code generation.
 Figure \ref{fig:ClassicalVision} shows the classical vision for Model-Driven Engineering.
@@
 Indeed, the model/code co-evolution problem remains an open subject in the software engineering research community and, consequently, software development projects often abandon the idea of maintaining the synchronization between model and source code during development process.
-In that context, the model is developed at the early stages of development process, used mainly to prototype software application.
+In that context, the model is developed at the early stages of development process, used mainly to prototype software applications.
 Then, it may be manually reviewed back at the end of development process, for documentation purposes.
@@
     \centering
-    \includegraphics[width=1.0 \columnwidth]{ClassicalVision.pdf}
+    \includegraphics[width=1.0\columnwidth]{ClassicalVision.pdf}
     \caption{Classical vision for Model-Driven Engineering}

### 2021-02-17 architecture.tex
Updates from Overleaf

@@
 
-The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
+The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, through the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
 

### 2021-02-17 pamelaProcess.tex
pamelaProcess after Antidote

@@
-We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle.
+We advocate for a strong coupling between model and sources code, to give architects and developers a way to both interact during the whole development cycle.
 PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization.
@@
     \includegraphics[width=1.0 \columnwidth]{PamelaVisionV2.pdf}
-    \caption{PAMELA approach for modelling}
+    \caption{PAMELA approach for modeling}
     \label{fig:PamelaVision}
@@
 This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations.
-The evolution rythm between models and code is low.
-This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rythm when the programmer also becomes the modeler.
-In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.
+The evolution rhythm between models and code is low.
+This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler.
+In fact, when the programmer identifies a pattern, an abstraction, a generalization, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.
 
@@
 The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms.
-They include consistency checkings, which constrain their use and helps the programmer to avoid inconsistencies or errors.
+They include consistency checking, which constrain their use and helps the programmer to avoid inconsistencies or errors.
 We have first experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.
 
-Our experience shows that introducing and reusing new concepts (1) reduce the size of the code, (2) reduce the risk of errors and (3) improve the code structure.
+Our experience shows that introducing and reusing new concepts (1) reduce the size of the code (2) reduce the risk of errors and (3) improve the code structure.
 The cycle of development between the model and the code can then be drastically reduced, leading to what we call \emph{continuous modeling}.
@@
 This scenario includes the case where the model (made by the modelers) is pre-existing.
-    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing this code by abstract method declarations in Java interface, reducing risks of errors.
+    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing code base (legacy) and refactor it while replacing this code by abstract method declarations in Java interface, reducing risks of errors.
     \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (e.g., Security Patterns \cite{silva20}) which offers code weaving at runtime, and runtime monitoring.

### 2021-02-17 architecture.tex
architecture after Antidote

--- 2021-02-16 architecture.tex
@@
 
-The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
+The PAMELA metamodel is depicted in Figure~\ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
 
@@
 PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities.
-Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods.
+Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API without any implementation for methods.
 Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
 will be responsible for providing custom code implementing domain or
-application specific behaviour. A partially implemented \texttt{abstract} Java
+application-specific behavior. A partially implemented \texttt{abstract} Java
 \texttt{class} may be defined as partial base implementation (conforms to
@@
 \begin{itemize}
-    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning desired value).
+    \item A \emph{read-only single property} will define read access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning desired value).
     \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking value to be set as unique argument).
-    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking value to be added as unique argument), a \emph{remover} (a Java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
+    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking value to be added as unique argument), a \emph{remover} (a Java method taking value to be removed as unique argument), and may define additional methods for extended features such as re-indexing, for example.
 \end{itemize}
@@
     \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
-    \item \texttt{@Reindexer(String)}: tag annotating method as unique reindexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
+    \item \texttt{@Reindexer(String)}: tag annotating method as unique re-indexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
     \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
@@
  \begin{itemize}
-     \item If the accessed method is part of a \emph{ModelProperty} (a getter, or a setter, etc..), and no custom implementation is defined neither in the class declared as implementation, nor in a class declared as partial implementation in the context of traits, then, execution is delegated to the related property implementation (generic code provided by the PAMELA interpreter).
+     \item If the accessed method is part of a \emph{ModelProperty} (a getter, or a setter, etc.), and no custom implementation is defined neither in the class declared as implementation, nor in a class declared as partial implementation in the context of traits, then, execution is delegated to the related property implementation (generic code provided by the PAMELA interpreter).
      \item If the accessed method is defined in a class declared as implementation, or in a class declared as partial implementation, then this method is executed.
@@
 
-This general scheme provides also an extension point allowing to instrument the code.
+This general scheme also provides an extension point allowing to instrument the code.
 This extension point is used in order to integrate other features such as notification management, undo/redo stack management, assertion checking at runtime (support for \emph{Design by Contract}, aka JML), and dynamic code weaving in the context of \emph{Aspect Programming}.
 
-PAMELA model at runtime is computed dynamically, working on the classpath of
+The PAMELA model at runtime is computed dynamically, working on the classpath of
 launched Java application, and starting from a simple Java interface (or a
@@
 (some already mentioned) such as: multiple inheritance and traits, containment
-management, cloning, fine-grained notifications management, object graph
+management, cloning, fine-grained notification management, object graph
 comparison and diff/merge, visiting patterns, clipboard management, validation,
@@
 Listing \ref{lst:model} shows a very basic model with two entities: \emph{Book} and \emph{Library}.
-Entity \emph{Book} defines two read-write single properties \emph{title} and \emph{ISBN} with single cardinality and with \texttt{String} type.
+Entity \emph{Book} defines two read-write single properties (\emph{title} and \emph{ISBN}) with single cardinality and with \texttt{String} type.
 Entity \emph{Book} also defines a constructor with initial \emph{title} value.
@@
 
-The lines 3--4 instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships.
+The lines 3--4 instantiate a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships.
 This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled bytecode.
@@
 
-Custom code can be easily added to this model as we show in Listing \ref{lst:custom}.
+Custom code can be easily added to this model as we show in Listing~\ref{lst:custom}.
 It shows how to integrate custom code to the fully interpreted \emph{Book} entity described above.
@@
 Custom implementations are defined using classical Java implementation/overrides scheme.
-Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, can not be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property.
-Note that this implementation references a default interpretated implementation (call to \texttt{performSuperGetter(String)} method).
+Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, cannot be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property.
+Note that this implementation references a default interpreted implementation (call to \texttt{performSuperGetter(String)} method).
 

### 2021-02-17 implementation.tex
implementation after Antidote

@@
 
-\subsection{Exposed API at design-time}
+\subsection{Exposed API at design time}
 
@@
 From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding.
-Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed.
-This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).
+Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed.
+This also provides an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at runtime (support for Design by Contract, aka JML).
 
@@
 
-Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.
+Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.
 
-A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developped to help measure that overhead.
+A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead.
 We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value).
 From this simple model, two java implementations have been derived using code generation.
-The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpretated (it uses PAMELA framework and defines only interfaces and API methods).
+The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses PAMELA framework and defines only interfaces and API methods).
 
-Unsurprisingly, we quantifiate a significant CPU overhead (from x20).
-This is due to the fact that \texttt{MethodHandler} mechanism is fully interpretated and contains many hooks, and cannot be compared to a fully-compiled and optimized Java dynamic binding.
-As well, memory footprint is increased from a x5 factor.
+Unsurprisingly, we quantify a significant CPU overhead (from $\times$20).
+This is due to the fact that \texttt{MethodHandler} mechanism is fully interpreted and contains many hooks, and cannot be compared to a fully compiled and optimized Java dynamic binding.
+As well, memory footprint is increased from an
+$\times$5 factor.
 
-These conclusions must be moderated by the fact that this overhead only applies to a very small portion of code (mostly property accessors methods).
+These conclusions must be moderated by the fact that this overhead only applies
+to a very small portion of the code (mostly property accessors methods).
 A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer.
-Using PAMELA won't increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case).
-As we will see in next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.
+Using PAMELA won’t increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case).
+As we will see in the next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.
 
-Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc...).
+Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).
 

### 2021-02-17 6-Experimentations.tex
remove orange comments (section 5.2 : Formose project)
experimentation after Antidote

@@
  The PAMELA framework has been successfully applied in a variety of complex programming and modeling scenarios and we continue to use it daily as part of our modeling toolbox.
-In the following, we describe three important uses cases in which PAMELA was a core component.
+In the following, we describe three important use cases in which PAMELA was a core component.
 
@@
 Federation)~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
-requests submissions for a standard addressing, ”federation of information
+requests submissions for a standard addressing ”federation of information
 across different representations, levels of abstraction, communities,
@@
 for model federation across multiple technological spaces. Conceptualization is
-addressed trough the proposition of a language called FML (Flexo Modeling
+addressed through the proposition of a language called FML (Flexo Modeling
 Language), which is executable on the platform. Openflexo infrastructure
@@
 various maturity stage regarding their industrialization (Microsoft Word, Excel
-and PowerPoint, EMF, OWL, Diagramming, JDBC, XML, OSLC, etc...)
+and PowerPoint, EMF, OWL, Diagraming, JDBC, XML, OSLC, etc.)
 
@@
 In most components, PAMELA framework is extensively used.
-As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagramming features) is composed of 998 classes.
-159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}.
+As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes.
+159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.
 
-The total base of code for Openflexo infrastructure represents around 900.000
-lines of Java code. Regarding backend modeling, the PAMELA framework is
-extensively used in Openflexo core as well as in most technology adapters, with
-very few specific implementations for properties. The observer/observable
-pattern is generally used for graphical user interfaces, which also rely on
-PAMELA framework.
+The total base of code for Openflexo infrastructure represents around
+\num{900000} lines of Java code. Regarding backend modeling, the PAMELA
+framework is extensively used in Openflexo core as well as in most technology
+adapters, with very few specific implementations for properties. The
+observer/observable pattern is generally used for graphical user interfaces,
+which also rely on PAMELA framework.
 
 An interesting experimentation has been done in the context of Openflexo development process.
-When Pamela has been integrated to the code base, a big portion of former legacy code has gradually and iteratively been migrated to Pamela.
+When Pamela has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to Pamela.
 Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration.
@@
 Pamela implementation is now really stable and mature.
-According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.
+According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.
 
@@
 
-The Formose ANR (french National Agency for Research) project
+The Formose ANR (French National Agency for Research) project
 (ANR-14-CE28-0009)\cite{FormoseWebSite} aimed to design a formally-grounded,
@@
 
-The main results of the project are a requirements modeling multi-views
-language, its associated design process and the development of an opensource
+The main results of the project are a requirements modeling multi-view
+language, its associated design process and the development of an open-source
 platform called Formod\cite{FormodWebSite}, built using Openflexo
-infrastructure and PAMELA framework. \todo{Say how and where PAMELA was used}
+infrastructure and PAMELA framework. %\todo{Say how and where PAMELA was used}
 The requirements modeling language is based on KAOS~\cite{kaos} for goal
@@
 
-\todo{Remark that it is industry ready}
+%\todo{Remark that it is industry ready}
 

### 2021-02-17 related.tex
related after Antidote

@@
-The PAMELA framework can be seen as a CASE tool that focuses in the design and verification and validation phases of the software development life-cycle.
+The PAMELA framework can be seen as a CASE tool that focuses on the design and verification and validation phases of the software development life-cycle.
 In that sense, it presents similarities to other existing object oriented CASE tools and/or language workbenches such as Kermeta~\cite{kermeta} and the Platypus~\cite{platypus} meta-case tool as they permit the definition and manipulation of models, metamodels and constraints/behavior.
-Two main features separate PAMELA from those approaches though: 1) The main artefact for the PAMELA framework is standard java code; and 2) PAMELA does not rely on code generation but on the interpretation/construction of models at runtime guided by code annotations.
+Two main features separate PAMELA from those approaches though: 1) The main artifact for the PAMELA framework is standard java code; and 2) PAMELA does not rely on code generation but on the interpretation/construction of models at runtime guided by code annotations.
 This allows for the seamlessly blending of the metamodeling/programming phases.
@@
 %However the focus of most models@runtime approaches is not on easing the development process but on enabling the analysis and/or the adaptation/evolution of ever running applications. 
-Examples of such approaches are: AC-Contrat~\cite{accontract} which provides runtime verification of properties for context-aware applications; Ramses~\cite{ramses} focused on dynamic adaptations; and \cite{denker2010modeling} that performs feature analysis.~\cite{song2010applying} wraps running systems in standard UML-like models in order to perform analysis and management tasks by using off-the-shelf MDE tools and techniques whereas FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime (Synchronization of changes between models and code are however limited for languages such as Java).
+Examples of such approaches are: AC-Contrat~\cite{accontract} which provides runtime verification of properties for context-aware applications; Ramses~\cite{ramses} focused on dynamic adaptations; and \cite{denker2010modeling} that performs feature analysis.~\cite{song2010applying} wraps running systems in standard UML-like models in order to perform analysis and management tasks by using off-the-shelf MDE tools and techniques whereas FAME \cite{kuhn2008fame} is a polyglot library that keeps metamodels accessible and adaptable at runtime (Synchronization of changes between models and code are, however, limited for languages such as Java).
 
@@
 
-More similar to PAMELA (as it focus on avoiding round-tripping issues) in~\cite{cavarle2016dynamic} the authors present an approach to keep the (bi-directional) synchronization between feature-based models and generated code alternatives.
+More similar to PAMELA (as it focus on avoiding round-tripping issues) in~\cite{cavarle2016dynamic} the authors present an approach to keep the (bidirectional) synchronization between feature-based models and generated code alternatives.
 They rely on Pharo\cite{pharo} reflective capabilities.

### 2021-02-17 architecture.tex

@@
 
-The PAMELA metamodel is depicted in Figure~\ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
+The PAMELA metamodel is depicted in Figure~\ref{fig:PamelaMetaModel}, and allows, through the use of annotations, the definition of \emph{simple} UML-like models directly on the code.
 

### 2021-02-17 ResponseLetter.tex
typo in response letter

@@
 }
- Answer to question X \todo{I do not know what to answer here.}
+We agree with the reviewer. The PAMELA documentation and web site have been updated in order to improve tests understanding (see section 2.2 of "Getting started"). 
 \end{response}
@@
 \begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
-We have remove this reference to PAMELA semantics in the paper. Our intention was to refer to the PAMELA metamodel.
+We have removed this reference to PAMELA semantics in the paper. Our intention was to refer to the PAMELA metamodel.
 \end{response}

### 2021-03-22 Todo-review2.tex
Added Todo-Review2.tex to keep track of changes

To trace what is corrected/or not

At the end : antidote...

Reviewer #1: The paper and presentation are much improved, and the software documentation improved somewhat, compared to the original submission.

There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:
- "to /avoid" on p3 (DONE)
- "detailed thereafter" -> "detailed hereafter" (It's Correct)
- "increasing PAMELA metamodel" -> "by _extending_ PAMELA_'s_ metamodel" (DONE)
- "with setter/getter": make both plural (DONE)
- POJO used here and elsewhere yet the acronym is introduced much later (DONE)
- p4 l11-12: reduce etc. should be reduce_s_ etc. (Keep as it is)
- odd spacing before : consistently throughout bullet list + text  (Keep as it is: I trust Latex)
- "in the section 3.2" -> "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them.
- "defines API": insert "an"
- "in Java interface": insert "a"
- "returning desired value", "taking value" etc.: insert "the"
- "May also declares"...
- p6 l46: @Embedded/"as embedded" meaning?
- p7 "of launched Java application" missing "the"
- "is/are" -> "is (are)", in line with parentheses earlier in sentence
- "Experimented programmers": I doubt it, unless you got major ethical clearance
- "Listing 2: model" -> "~: Model" consistent with Listing 1
- the book title is "Lord of the ring_s_" with s
- "using classical Java ... scheme": missing "the"
- "Java programming language": insert "The"
- l43 inconsistent spacing
- l46 package ... package
- p11 l36 sticking into margin
- S.4.3 please add size in LoC and classes
- java -> Java
- p13 "diana component": insert "the"
- "experimentation" -> "experiment"
- "When Pamela has been" -> "After Pamela was"
- l36 the quotation mark before code generation is the wrong one
- p 14 "appertaining" -> "pertaining", "relating"

Reviewer #2: This new version is really better than the first one.

My major concern is related to section 2. 
- Would it be possible to structure it more, because currently it is several paragraphs the one bellow the others.
- figure 2 is not enough described in section 2 and even in the paper.

More over, concerning section 4.4, I appreciate this new section. However, while reading, it is not a good advertisement for Pamela. Overhead (x20) is huge. Perhaps it is not the right benchmark if it provides bad results that are not representing the right usage.

Minors:
- POJO should be explain the first time p3 not the second time.
- p5 it is written twice that ModelEntities are encoded in a Java Interface
- p13, it is written twice that Pamela framework is extensively used.

### 2021-03-22 architecture.tex
Added Todo-Review2.tex to keep track of changes

@@
-PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one runtime component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.
+PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one runtime component (the PAMELA interpreter), described in Section \ref{sub:RunTime}.
 

### 2021-03-22 pamelaProcess.tex
Added Todo-Review2.tex to keep track of changes

@@
 Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted).
-The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.
+The idea is to avoid separation between modeling and code to facilitate consistency management and to avoid round-trip issues.
 Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.
@@
 This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler.
-In fact, when the programmer identifies a pattern, an abstraction, a generalization, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.
+In fact, when the programmer identifies a pattern, an abstraction, a generalization, s/he can use PAMELA to develop and capitalize on this abstraction by extending PAMELA's metamodel.
 
@@
 They include consistency checking, which constrain their use and helps the programmer to avoid inconsistencies or errors.
-We have first experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.
+We have first experimented their use with setters/getters to define Plain Old Java Objects (POJO), with traits to implement multiple inheritance or with roles and rules to set security rules on classes.
 

### 2021-03-23 6-Experimentations.tex
review up to p5 (details only)

@@
 The full Openflexo infrastructure is composed of about 50 components.
-In most components, PAMELA framework is extensively used.
+In most components, the PAMELA framework is extensively used.
 As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes.
@@
 observer/observable pattern is generally used for graphical user interfaces,
-which also rely on PAMELA framework.
+which also rely on the PAMELA framework.
 
 An interesting experimentation has been done in the context of Openflexo development process.
-When Pamela has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to Pamela.
+When PAMELA has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to PAMELA.
 Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration.
@@
 
-Pamela implementation is now really stable and mature.
-According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.
+The PAMELA implementation is now really stable and mature.
+According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no PAMELA-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.
 
@@
 platform called Formod\cite{FormodWebSite}, built using Openflexo
-infrastructure and PAMELA framework. %\todo{Say how and where PAMELA was used}
+infrastructure and the PAMELA framework. %\todo{Say how and where PAMELA was used}
 The requirements modeling language is based on KAOS~\cite{kaos} for goal

### 2021-03-23 Todo-review2.tex
review up to p5 (details only)

@@
 - "in the section 3.2" -> "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
-- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them.
-- "defines API": insert "an"
-- "in Java interface": insert "a"
-- "returning desired value", "taking value" etc.: insert "the"
-- "May also declares"...
+- p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them. (Done ; with Pamela -> PAMELA ; to be checked)
+- "defines API": insert "an" (DONE)
+- "in Java interface": insert "a" (DONE)
+- "returning desired value", "taking value" etc.: insert "the" (DONE)
+- "May also declares"... (DONE)
 - p6 l46: @Embedded/"as embedded" meaning?
@@
 Minors:
-- POJO should be explain the first time p3 not the second time.
-- p5 it is written twice that ModelEntities are encoded in a Java Interface
+- POJO should be explain the first time p3 not the second time. (DONE)
+- p5 it is written twice that ModelEntities are encoded in a Java Interface (DONE)
 - p13, it is written twice that Pamela framework is extensively used.

### 2021-03-23 architecture.tex
review up to p5 (details only)

@@
 
-A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}.
-A \texttt{ModelEntity} reflects a concept and is encoded in a Java \texttt{interface}.
-PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities.
-Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API without any implementation for methods.
+A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{Model\-Entity} reflects a concept and is encoded in a Java \texttt{interface}. 
+Then, a \texttt{Model\-Entity} only defines an API (Application Programming Interface) without any implementation for methods. 
 Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
@@
 (since Java 8).
+The PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities.
 
@@
 \begin{itemize}
-    \item A \emph{read-only single property} will define read access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning desired value).
-    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking value to be set as unique argument).
-    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking value to be added as unique argument), a \emph{remover} (a Java method taking value to be removed as unique argument), and may define additional methods for extended features such as re-indexing, for example.
+    \item A \emph{read-only single property} will define read access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning the desired value).
+    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking the value to be set as unique argument).
+    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking the value to be added as unique argument), a \emph{remover} (a Java method taking the value to be removed as unique argument), and may define additional methods for extended features such as re-indexing, for example.
 \end{itemize}
@@
 \begin{itemize}
-    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declare an abstract entity.
+    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May also declare an abstract entity.
     \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.

### 2021-03-23 implementation.tex
review up to p5 (details only)

@@
 
-From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding.
+From a technical point of view, the PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding.
 Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed.
@@
 
-PAMELA implementation is modularized.
+The PAMELA implementation is modularized.
 
@@
 From this simple model, two java implementations have been derived using code generation.
-The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses PAMELA framework and defines only interfaces and API methods).
+The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).
 
@@
 
-Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).
+Moreover, comparison between the two implementations should be tempered by the fact that the PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).
 

### 2021-03-23 6-Experimentations.tex
review up to p13 (but the quotation mark)

@@
 In most components, the PAMELA framework is extensively used.
-As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes.
+As an example the \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes.
 159 of those classes (mostly the diagraming model) are defined as Pamela \texttt{ModelEntity}.
@@
 
-An interesting experimentation has been done in the context of Openflexo development process.
-When PAMELA has been integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to PAMELA.
+An interesting experiment has been done in the context of Openflexo development process.
+When PAMELA was integrated to the code base, a big portion of the former legacy code has gradually and iteratively been migrated to PAMELA.
 Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration.

### 2021-03-23 Todo-review2.tex
review up to p13 (but the quotation mark)

@@
 - "to /avoid" on p3 (DONE)
-- "detailed thereafter" -> "detailed hereafter" (It's Correct)
+- "detailed thereafter" -> "detailed hereafter" (It's Correct : As adverbs the difference between hereafter and thereafter is that hereafter is in time to come; in some future time or state while thereafter is after that, from then on.)
 - "increasing PAMELA metamodel" -> "by _extending_ PAMELA_'s_ metamodel" (DONE)
@@
 - p6 l46: @Embedded/"as embedded" meaning?
-- p7 "of launched Java application" missing "the"
-- "is/are" -> "is (are)", in line with parentheses earlier in sentence
-- "Experimented programmers": I doubt it, unless you got major ethical clearance
-- "Listing 2: model" -> "~: Model" consistent with Listing 1
-- the book title is "Lord of the ring_s_" with s
-- "using classical Java ... scheme": missing "the"
-- "Java programming language": insert "The"
-- l43 inconsistent spacing
-- l46 package ... package
-- p11 l36 sticking into margin
-- S.4.3 please add size in LoC and classes
-- java -> Java
-- p13 "diana component": insert "the"
-- "experimentation" -> "experiment"
-- "When Pamela has been" -> "After Pamela was"
+- p7 "of launched Java application" missing "the" (DONE)
+- "is/are" -> "is (are)", in line with parentheses earlier in sentence (DONE)
+- "Experimented programmers": I doubt it, unless you got major ethical clearance (Replaced by "Experienced PAMELA programmers")
+- "Listing 2: model" -> "~: Model" consistent with Listing 1 (DONE)
+- the book title is "Lord of the ring_s_" with s (DONE)
+- "using classical Java ... scheme": missing "the" (Check :mis au pluriel)
+- "Java programming language": insert "The" (DONE)
+- l43 inconsistent spacing (DONE)
+- l46 package ... package (DONE)
+- p11 l36 sticking into margin (Final check to do)
+- S.4.3 please add size in LoC and classes (?)
+- java -> Java (DONE)
+- p13 "diana component": insert "the" (DONE)
+- "experimentation" -> "experiment" (DONE)
+- "When Pamela has been" -> "After Pamela was" (DONE : if we consider it's over)
 - l36 the quotation mark before code generation is the wrong one

### 2021-03-23 architecture.tex
review up to p13 (but the quotation mark)

@@
     \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as
-      embedded. %according to PAMELA semantics.
+      embedded \todo{meaning?}. %according to PAMELA semantics.
     \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
@@
 The PAMELA model at runtime is computed dynamically, working on the classpath of
-launched Java application, and starting from a simple Java interface (or a
-collection of Java interfaces) which is/are PAMELA-annotated. From a
+the launched Java application, and starting from a simple Java interface (or a
+collection of Java interfaces) which is (are) PAMELA-annotated. From a
 mathematical point of view, internal representation of the underlying model is
@@
 official web site\footnote{\url{https://pamela.openflexo.org}}.
-Experimented programmers may extend PAMELA defining new annotations.
+Experienced PAMELA programmers may extend PAMELA defining new annotations.
 
@@
 %    \centering
-\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize, caption=model execution/manipulation, label=lst:execution]
+\begin{lstlisting}[language=Java,basicstyle=\ttfamily\footnotesize, caption=Model execution/manipulation, label=lst:execution]
 // Instantiate the meta-model
@@
 Book myFirstBook 
-    = factory.newInstance(Book.class, "Lord of the ring");
+    = factory.newInstance(Book.class, "Lord of the rings");
 Book anOtherBook = factory.newInstance(Book.class, "Holy bible");
@@
 The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity.
-Custom implementations are defined using classical Java implementation/overrides scheme.
+Custom implementations are defined using classical Java implementation/override schemes.
 Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, cannot be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property.

### 2021-03-23 implementation.tex
review up to p13 (but the quotation mark)

@@
 Such a language supports the attribute-oriented programming if its grammar allows adding custom declarative tags to annotate standard program elements.
-Java programming language is a good candidate, as it supports annotations.
+The Java programming language is a good candidate, as it supports annotations.
 
-The PAMELA API exposed to the developer mainly consists of : 1) a set of annotations; and 2) a set of unimplemented Java interfaces exposing required features.
+The PAMELA API exposed to the developer mainly consists of: 1) a set of annotations; and 2) a set of unimplemented Java interfaces exposing required features.
 
-The package \texttt{org.openflexo.pamela.annotations} package exposes the set of annotations which were presented in Subsection \ref{sub:DesignTime}.
+The \texttt{org.openflexo.pamela.annotations} package exposes the set of annotations which were presented in Subsection \ref{sub:DesignTime}.
 
-The package \texttt{org.openflexo.pamela} contains the following feature-related Java interfaces:
+The \texttt{org.openflexo.pamela} package contains the following feature-related Java interfaces:
 \begin{itemize}
@@
 
-Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.
+Compared to a basic POJO implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.
 
@@
 We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value).
-From this simple model, two java implementations have been derived using code generation.
-The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).
+From this simple model, two Java implementations have been derived using code generation.
+The first implementation was made using fully implemented Java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).
 

### 2021-03-23 Todo-review2.tex
small corrections completed. Remain: improve fig2 comments, restructure section 2, rephrase 4.4

@@
 - the book title is "Lord of the ring_s_" with s (DONE)
-- "using classical Java ... scheme": missing "the" (Check :mis au pluriel)
+- "using classical Java ... scheme": missing "the" (To check :mis au pluriel)
 - "Java programming language": insert "The" (DONE)
@@
 - "When Pamela has been" -> "After Pamela was" (DONE : if we consider it's over)
-- l36 the quotation mark before code generation is the wrong one
-- p 14 "appertaining" -> "pertaining", "relating"
+- l36 the quotation mark before code generation is the wrong one 'changed to ' and ' ?)
+- p 14 "appertaining" -> "pertaining", "relating" (Kept as is.
+As verbs the difference between pertain and appertain is that pertain is to belong while appertain is to belong to or be a part of, whether by right, nature, appointment, or custom; to relate to.)
 
@@
 - p5 it is written twice that ModelEntities are encoded in a Java Interface (DONE)
-- p13, it is written twice that Pamela framework is extensively used.
+- p13, it is written twice that Pamela framework is extensively used. (DONE : 1 changed to largely)

### 2021-04-21 ResponseLetterMinor.tex
Adding minor revision response letter

--- 2021-02-15 ResponseLetter.tex
@@
-% Title:    A LaTeX Template For Responses To a Referees' Reports
+ Title:    A LaTeX Template For Responses To a Referees' Reports
 % Author:   Petr Zemek <s3rvac@gmail.com>
@@
 
-\title{Response letter to Manuscript SCICO-D-20-00176 entitled "PAMELA: an Annotation-Based Java Modeling Framework"}
+\title{Response letter to Manuscript SCICO-D-20-00176R1 entitled "PAMELA: an Annotation-Based Java Modeling Framework"}
 %\author{Author1 \and Author2 \and Author3}
@@
 \bigskip
-Thanks for your message of January 9th, 2021, in which you suggested preparing a major revision of our manuscript  \textbf{PAMELA: an Annotation-Based Java Modeling Framework}.
+Thanks for your message of Mars 15th, 2021, in which you suggested preparing a minor revision of our manuscript  \textbf{PAMELA: an Annotation-Based Java Modeling Framework}.
 
@@
 We are grateful to the reviewers for their interest, effort, and suggestions.
-We detail in the following pages how we have addressed the reviewers' concerns; we are confident that the paper is now stronger, and hope that you will agree.
-We have organized the responses to the comments on a per-reviewer basis, focusing, as suggested, on:
+We detail in the following pages how we have addressed the reviewers' concerns;  We have organized the responses to the comments on a per-reviewer basis, focusing, as suggested, on:
 
 \begin{itemize}
-\item Improving the readability of the manuscript. 
-\item Improving tool documentation and examples.
-\item Provide additional experiments \& evaluation.
+\item Fixing typos, language errors, and other minor issues.
+\item Refactoring Section 2.
+\item Provide further explanations with respect to the performance of the approach.
 \end{itemize}
@@
 \bigskip
-Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador and Antoine
+Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador, Fabien and Antoine
 
@@
 \Reviewer{\#1}
-\begin{response}{They then briefly discuss implementation details, in a single page; and then discuss 'industrial use cases and experiments'.
-
-Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
-} We agree with the reviewer.
-We have extended the manuscript in order to provide more details about PAMELA itself, its usage and its performance.
-Concretely, we have added code base metrics in Section 4.3., a performance evaluation and discussion in Section 4.4, and more details on the role PAMELA took in the industrial use cases and experiments described in Section 5.
-\end{response}
-
-\begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
-We fully agree with the reviewer.
-We have expanded the related work section and compared our work with existing metamodeling and models@runtime approaches.
-Moreover, we have included a number of new references as suggested by Reviewer 2.
-\end{response}
-
-\begin{response}{What would also be interesting is a general discussion on performance of your tool, gains by using your tool, ...
-} We discuss PAMELA performance in a new Section 4.4.
-We have created a workbench consisting in an small model (4 entities containing 4-5 properties) and measured how well PAMELA fares against a basic POJO implementation when dealing with large instance models.
-\end{response}
-
-\begin{response}{Your reference list is incomplete and inconsistent.
-Formod or Formosa?
-When were URLs accessed (add date).
-DOIs are missing.
-Ref.
-5 should have the ISoLA track name and editors added.
-Ref.
-7 is missing publisher, journal, conference.
-Inconsistent use of first names, first names + initials, first initial.}
-We have expanded our reference list and fixed all (detected) insconsistencies.
-\end{response}
-
-\begin{response}{The paper highlights include "Contract programming with assertions [remove the plural] checking at runtime", yet that is not mentioned in the paper at all.
-} Thank you for pointing this out. We have removed this point from the highlights as it is not on the focus of the paper.
-\end{response}
-
-\begin{response}{That also brings up the question: what about correctness of your code? Any evidence for that?}
-PAMELA have been thoroughly tested.
-The core PAMELA implementation is accompained with a battery of unit tests (6k lines of code and 112 classes) reaching a coverage of 66\%.
-We have modified the paper in order to add this information in a new Section 4.3.
-\end{response}
-
-\begin{response}{p10 l32+: very little detail or quantification on the tool's code base.
-Surely you can say more about its structure, size of components, ...
-other interesting and relevant observations on the codebase?}
-The new Section 4.3. provides details about the PAMELA code base.
-\end{response}
-
-\begin{response}{Detailed comments}
-We have fixed all the issued highlighted in the detailed comments.
-\end{response}
-
-\ToolComments{\#1}
-
-\begin{response}{gives "git clone git@github.com:openflexo-team/pamela.git", which (even when using my own username) does not work.
-However, "git clone https://github.com/openflexo-team/pamela.git" immediately does.
-Why not state the latter?
-}
-
-We apologize for the inconvenience. We have replaced "git clone git@github.com:openflexo-team/pamela.git" for "git clone https://github.com/openflexo-team/pamela.git" as suggested.
-\end{response}
-
-\begin{response}{how to actually build the environment is not stated for this option.}
-We apologize for the inconvenience. The PAMELA documentation and web site have been updated in order to solve the problem.
-\end{response}
-
-\begin{response}{Run tests Unit tests located in src/test/java in pamela-core are a good starting point to discover the framework." No further information is given.
-Apparently, the user has to figure out him/herself how to run the tests.
-Documentation is lacking here.
-}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{Make your first model." And then what?
-}
-We have refactored the PAMELA documentation on the web site so that it redirects to the getting started guide.
-\end{response}
-
-\begin{response}{Summarizing: the *getting started* documentation is incomplete and certainly not suitable for getting started.
-The page "https://pamela.openflexo.org/SNAPSHOT/example.html" gives a basic example, but the relation between that and the getting started page is not made explicit, nor do the two in their current shape provide a user with a proper getting started, allowing him/her to at least (1) install the framework, (2) create a basic example, (3) use the framework with that basic example.
-This needs to be improved before an evaluation of the software can be completed.
-}
-We apologize again for the lack of clarity in the documentation.
-It has been refactored so that the starting point is clear and the reader is guided step by step to the realization/usage of a first model.
+\begin{response}{There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:} 
+All highlighted typos, languages errors and minor issues have been fixed.
 \end{response}
@@
 \Reviewer{\#2}
-\begin{response}{scenarios are missing. As a developer I don't know in which cases it is relevant to use this tool.}
-We have rewritten Section 2 so that it includes now a list and brief description of typical usage scenarios for the PAMELA framework (e.g., coding-metamodeling blending, aspect-oriented programming or legacy code re-engineering).
-We also state that the standard PAMELA user will reuse existing annotation while more advanced users may become (meta) modelers and provide their own abstractions in a continous modeling approach.
+\begin{response}{Would it be possible to structure it more, because currently it is several paragraphs the one bellow the others.}
+We will probably do something...
 \end{response}
 
-\begin{response}{how to use the tool is not describe properly.
-It is not clear if the developers need to write all the annotations.
-If an application is built with PAMELA, what is the link between the interfaces and the concrete classes.
-Indeed, at one point, code corresponding to the behaviour will be needed; and not only the accessors to navigate the structure of the model.}
-We have modified Section 2 to clarify that standard PAMELA users will reuse existing annotations which are accompanied with default implementations in the PAMELA interpreter.
-Additional details are given in Sections 3 and 4 (and in the online documentation: \url{https://pamela.openflexo.org/SNAPSHOT/pamela-core/3-metaprogramming_support.html}) See Listing 3 for an example of how the default implementation can be replaced by custom ones.
+\begin{response}{figure 2 is not enough described in section 2 and even in the paper.}
+We have removed figure 2 ;) haha!
 \end{response}
 
-\begin{response}{how does such "code" evolve with time? Authors have used PAMELA for years, how such applications resist with time?}
-PAMELA helps factorizing code corresponding to recurrent concepts.
-In that sense its use does reduce the size and duplication of code, decreases the number of bugs and improves the code structure.
-All this makes final applications that use PAMELA (more) easily maintainable.
-We briefly discuss in Section 5.1. how integrating PAMELA in the OpenFlexo tool helped removing duplicated code and eliminating bugs in the latter.
+\begin{response}{More over, concerning section 4.4, I appreciate this new section.
+However, while reading, it is not a good advertisement for Pamela.
+Overhead (x20) is huge.
+Perhaps it is not the right benchmark if it provides bad results that are not representing the right usage.}
+We totally agree...
 \end{response}
 
-\begin{response}{section 2 is too abstract to be very useful}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
-We have remove this reference to PAMELA semantics in the paper. Our intention was to refer to the PAMELA metamodel.
-\end{response}
-
-\begin{response}{the partially implemented class concept is not clear.}
-The partially implemented class concept is very similar to the default methods introduced in Java 8 (which were not available when PAMELA was first created) permitting a sort of multiple inheritance.
-We have clarified this in the paper.
-\end{response}
-
-\begin{response}{the additional features seem relevant, but they are not described. Either the authors describe them a bit more or they remove the section.}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{in section 3.4 it is not clear who writes what and at which step of the development process.}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{in section 5, the use of PAMELA in the two projects is not clear.}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{references are missing. For example Kermeta done, Platypus (from Alain Plantec), work of Glenn Cavarle and others from the model @runtime community.}
-We have extended the related work section in order to discuss the suggested related works (among others).
-\end{response}
-
-\begin{response}{in section 6, the authors write: "PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviours; and 2) maintaining a continuous cycle of metamodeling...
-These two points are not clear previously in the paper.
-}
-We have removed that part as it was indeed misleading and not discussed in that form previously in the paper. We have rephrased as follows:
-
-PAMELA is not focused on adapting or observing ever running programs (although it can be used for the verification of runtime properties as demonstrated in [20]) but on:  1) providing a mechanism to blend coding and metamodeling so that the code and the metamodel  may  be  built  incrementally  without  the  need  for  code  generation and thus,  avoiding round-tripping issues;  and 2) providing default implementations for frequently used abstractions in order to ease development.
-\end{response}
-
-\ToolComments{\#2}
-
-\begin{response}{The link in the paper is not clickable and copy-and-paste does not work (because of latex?) Link found: https://github.com/openflexo-team/pamela}
-Thanks for pointing this out. We have modified the latex settings so that this link (and all the others throughout the paper) is now clickable.
-\end{response}
-
-\begin{response}{Installation
-The readme of the GitHub page should link to the installation page found in https://pamela.openflexo.org. It is unclear we must visit this web site to install the tool.
-}
-  Response to n-ieme comment.
-\end{response}
-
-\begin{response}{
-What I have done
-
-Create a new project in Intellij Idea with gradle
-Add the line found in \url{https://pamela.openflexo.org/SNAPSHOT/getting_started.html}
-
-- Test 1: Create a model with only one entity without error
-
-- Test 2: Create the example of the paper
-
---->Bug: in Listing2, the authors execute Line9 ".setName" but the method does not exists
-Else, the example works.
-
-}
-  Response to n-ieme comment.
+\begin{response}{Minors}
+We have fixed all the highlighted minor issues.
 \end{response}

### 2021-04-21 1-Main.tex
fix compilation warnings
typos
Adding response letter template
ad package siunitx
Updates from Overleaf
Adding minor revision response letter

@@
 \usepackage{xcolor}
+\usepackage{siunitx}
+\usepackage{tikz-uml}
 \usepackage{hyperref}
-\usepackage{tikz-uml}
+\hypersetup{pdfauthor={Sylvain Guérin, Jean-Christophe Bach, Antoine Beugnard, Joel Champeau, Salvador Mart\'inez, Guillaume Pollet, Caine Silva}}
 
@@
 \newcommand{\ab}[1]{{\color{blue} #1}} % draft mode
+\newcommand{\jcb}[1]{{\color{red} #1}} % draft mode
+\newcommand{\sg}[1]{{\color{purple} #1}} % draft mode
 
@@
 
+\author[mysecondaryaddress]{Fabien Dagnat}
+%\ead{firstname.lastname@imt-atlantique.fr}
+
 \author[mysecondaryaddress]{Antoine Beugnard}
@@
 \address[mymainaddress]{ENSTA Bretagne, Lab-STICC, UMR 6285, Brest, France}%vérifier si c'est l'affiliation officielle
-\address[mysecondaryaddress]{IMT Atlantique, Lab-STICC, UMR 6285, F-29238 Brest, France}
+\address[mysecondaryaddress]{IMT Atlantique, Lab-STICC, UMR 6285, Brest, France}
 

### 2021-04-21 introduction.tex
Adding minor revision response letter

@@
 Conversely, we propose a shift in the modeling paradigm in which models and code are developed together and at the same time in what we call a continuous modeling process.
-The PAMELA framework supports this paradigm shift by providing the means to: 1) annotate Java code with model-based annotations; and 2) interpret these annotations at runtime.
+The PAMELA framework supports this paradigm shift by providing the means to:
+% Joel 1) annotate Java code with model-based annotations; 
+%and 2) interpret these annotations at runtime.
+1) weave model-based annotations with Java source code
+and 2) interpret model annotations at run-time.
 

### 2021-04-21 ResponseLetterMinor.tex
restructuration section 2

@@
- Title:    A LaTeX Template For Responses To a Referees' Reports
+% Title:    A LaTeX Template For Responses To a Referees' Reports
 % Author:   Petr Zemek <s3rvac@gmail.com>
@@
 \begin{response}{There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:}
+
 All highlighted typos, languages errors and minor issues have been fixed.
+
+\begin{itemize}
+
+\item "to /avoid" on p3 (DONE)
+\item "detailed thereafter" $->$ "detailed hereafter" (It's Correct : As adverbs the difference between hereafter and thereafter is that hereafter is in time to come; in some future time or state while thereafter is after that, from then on.)
+\item "increasing PAMELA metamodel" $->$ "by \_extending\_ PAMELA\_'s\_ metamodel" (DONE)
+\item "with setter/getter": make both plural (DONE)
+\item POJO used here and elsewhere yet the acronym is introduced much later (DONE)
+\item p4 l11-12: reduce etc. should be reduce\_s\_ etc. (Keep as it is)
+\item odd spacing before : consistently throughout bullet list + text  (Keep as it is: trust Latex)
+\item "in the section 3.2" $->$ "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
+\item p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them. (Done ; with Pamela $->$ PAMELA)
+\item "defines API": insert "an" (DONE)
+\item "in Java interface": insert "a" (DONE)
+\item "returning desired value", "taking value" etc.: insert "the" (DONE)
+\item "May also declares"... (DONE)
+\item p6 l46: @Embedded/"as embedded" meaning?
+\item p7 "of launched Java application" missing "the" (DONE)
+\item "is/are" $->$ "is (are)", in line with parentheses earlier in sentence (DONE)
+\item "Experimented programmers": I doubt it, unless you got major ethical clearance (Replaced by "Experienced PAMELA programmers")
+\item "Listing 2: model" $->$ "~: Model" consistent with Listing 1 (DONE)
+\item the book title is "Lord of the ring\_s\_" with s (DONE)
+\item "using classical Java ... scheme": missing "the" (DONE)
+\item "Java programming language": insert "The" (DONE)
+\item l43 inconsistent spacing (DONE)
+\item l46 package ... package (DONE)
+\item p11 l36 sticking into margin (Final check to do)
+\item S.4.3 please add size in LoC and classes (?)
+\item java $->$ Java (DONE)
+\item p13 "diana component": insert "the" (DONE)
+\item "experimentation" $->$ "experiment" (DONE)
+\item "When Pamela has been" $->$ "After Pamela was" (DONE : if we consider it's over)
+\item l36 the quotation mark before code generation is the wrong one 'changed to ' and ' ?)
+\item p 14 "appertaining" $->$ "pertaining", "relating" (Kept as is.
+As verbs the difference between pertain and appertain is that pertain is to belong while appertain is to belong to or be a part of, whether by right, nature, appointment, or custom; to relate to.)
+\end{itemize}
+
 \end{response}
@@
 \begin{response}{Would it be possible to structure it more, because currently it is several paragraphs the one bellow the others.}
-We will probably do something...
+
+Section 2 has been decomposed into 2 subsections
+
 \end{response}
@@
 \begin{response}{figure 2 is not enough described in section 2 and even in the paper.}
-We have removed figure 2 ;) haha!
+
+The end of 2.1 subsection provides an explanation of figure 2.
+
 \end{response}
@@
 Perhaps it is not the right benchmark if it provides bad results that are not representing the right usage.}
-We totally agree...
+
+SM comments
+
 \end{response}

### 2021-04-21 pamelaProcess.tex
Adding minor revision response letter
restructuration section 2

@@
 We advocate for a strong coupling between model and sources code, to give architects and developers a way to both interact during the whole development cycle.
+
+\subsection{Architecture overview}
+
 PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization.
@@
 The idea is to avoid separation between modeling and code to facilitate consistency management and to avoid round-trip issues.
-Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.
+Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.
+The left side shows the structure of the application source code where java files contain the code and annotations that link part of the code to PAMELA model entities such as class, attribute, method, etc.
+The right side shows that at runtime the PAMELA interpreter maintains the relationships between the application binary, result of the java compilation, and the internal representation of PAMELA models.
+The preservation of the structure and of the whole information allow a very good and high level of control over the execution.
+In the following section, we present some indications on the different ways of programming using PAMELA.
 
@@
 %\subsection{PAMELA Modeling/Development Process}
+\subsection{Usage of PAMELA}
 
@@
 This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations.
+% Joel : une ref pour Jakarta ?
 The evolution rhythm between models and code is low.
@@
 In fact, when the programmer identifies a pattern, an abstraction, a generalization, s/he can use PAMELA to develop and capitalize on this abstraction by extending PAMELA's metamodel.
-
+% Joel : le MM de Pamela n'est pas étendu en tant que tel (=M3) c'est le niveau M2 qui étendu ??? mais la bibliothèque de metamodèles est étendue ??? donc juste mettre PAMELA's metamodels. ????
 % Je pense qu'ici il faudrait plutot parler du modèle. Le métamodèle PAMELA est celui représenté figure 3. Ou alors parler du metametamodele PAMELA
@@
 The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms.
-They include consistency checking, which constrain their use and helps the programmer to avoid inconsistencies or errors.
+They include consistency checking, which constrain their use and help the programmer to avoid inconsistencies or errors.
 We have first experimented their use with setters/getters to define Plain Old Java Objects (POJO), with traits to implement multiple inheritance or with roles and rules to set security rules on classes.

### 2021-04-22 ResponseLetterMinor.tex
commented all minor issues details in response letter 2

--- 2021-04-21 ResponseLetterMinor.tex
@@
 \begin{response}{There remain or are introduced however, a substantial number of typos, language errors, and other minor issues that need correcting:}
-All highlighted typos, languages errors and minor issues have been fixed.
+
+All highlighted typos, languages errors and minor issues have been checked and fixed.
+
+%\begin{itemize}
+
+%\item "to /avoid" on p3 (DONE)
+%\item "detailed thereafter" $->$ "detailed hereafter" (It's Correct : As adverbs the difference between hereafter and thereafter is that hereafter is in time to come; in some future time or state while thereafter is after that, from then on.)
+%\item "increasing PAMELA metamodel" $->$ "by \_extending\_ PAMELA\_'s\_ metamodel" (DONE)
+%\item "with setter/getter": make both plural (DONE)
+%\item POJO used here and elsewhere yet the acronym is introduced much later (DONE)
+%\item p4 l11-12: reduce etc. should be reduce\_s\_ etc. (Keep as it is)
+%\item odd spacing before : consistently throughout bullet list + text  (Keep as it is: trust Latex)
+%\item "in the section 3.2" $->$ "in Section 3.2" (style: named references deserve a capital and no article before) (DONE)
+%\item p5 and many times: "PAMELA metamodel", "PAMELA model", "PAMELA framework" need "the" before them. (Done ; with Pamela $->$ PAMELA)
+%\item "defines API": insert "an" (DONE)
+%\item "in Java interface": insert "a" (DONE)
+%\item "returning desired value", "taking value" etc.: insert "the" (DONE)
+%\item "May also declares"... (DONE)
+%\item p6 l46: @Embedded/"as embedded" meaning?
+%\item p7 "of launched Java application" missing "the" (DONE)
+%\item "is/are" $->$ "is (are)", in line with parentheses earlier in sentence (DONE)
+%\item "Experimented programmers": I doubt it, unless you got major ethical clearance (Replaced by "Experienced PAMELA programmers")
+%\item "Listing 2: model" $->$ "~: Model" consistent with Listing 1 (DONE)
+%\item the book title is "Lord of the ring\_s\_" with s (DONE)
+%\item "using classical Java ... scheme": missing "the" (DONE)
+%\item "Java programming language": insert "The" (DONE)
+%\item l43 inconsistent spacing (DONE)
+%\item l46 package ... package (DONE)
+%\item p11 l36 sticking into margin (Final check to do)
+%\item S.4.3 please add size in LoC and classes (?)
+%\item java $->$ Java (DONE)
+%\item p13 "diana component": insert "the" (DONE)
+%\item "experimentation" $->$ "experiment" (DONE)
+%\item "When Pamela has been" $->$ "After Pamela was" (DONE : if we consider it's over)
+%\item l36 the quotation mark before code generation is the wrong one 'changed to ' and ' ?)
+%\item p 14 "appertaining" $->$ "pertaining", "relating" (Kept as is.
+As verbs the difference between pertain and appertain is that pertain is to belong while appertain is to belong to or be a part of, whether by right, nature, appointment, or custom; to relate to.)
+%\end{itemize}
+
 \end{response}
 
-\pagebreak
+%\pagebreak
 

### 2021-04-22 ResponseLetterMinor.tex
typos in response letter 2

@@
 
-Section 2 has been decomposed into 2 subsections
+Section 2 has been decomposed into two Subsections: Subsection 2.1.  Architecture overview
+and Subsection 2.2.  Usage of PAMELA.
 
@@
 
-The end of 2.1 subsection provides an explanation of figure 2.
+The end of Subsection 2.1 provides now a full explanation of figure 2.
 
@@
 
-SM comments
+We agree with the reviewer that the overhead may seem huge.
+Note, however, that these measures correspond to the very worst-case scenario: an application with absolutely no business logic.
+Although such applications do not correspond to the normal usage, we believe showing this measurement is important to fully characterize the performance of the approach.
+E.g., PAMELA should not be used when the application is a mere data structure.
+
+For normal applications, the time passed on attribute accessors is negligible compared to the time spent on business logic.
+We have verified this by using a profiling tool on PAMELA based applications (we have added this information to the paper).
+Moreover, as we show in Section 5 PAMELA has been successfully applied to many industrial projects, with no performance issues reported.
 

### 2021-04-22 implementation.tex
small corrections completed. Remain: improve fig2 comments, restructure section 2, rephrase 4.4
Adding minor revision response letter
restructuration section 2
typos in response letter 2

@@
 \end{itemize}
+% Joel : tu fais toujours référence à JML mais tu pourrais juste mentionner le Contract management ou Design By Contract approach ??
 
@@
 \begin{itemize}
-    \vspace{-0.2cm}\item Core implementation (\texttt{pamela-core}) provides all base features for PAMELA.
-It contains 20k lines of code involving 184 classes.
-This code is covered with unit tests (6k lines of code, 112 classes).
-Reached code coverage is about 66\%.
+    \vspace{-0.2cm}\item Core implementation (\texttt{pamela-core}) provides all base features. %for PAMELA. 
+    It contains 20k lines of code involving 184 classes. This code is covered with unit tests (6k lines of code, 112 classes). Reached code coverage is about 66\%.
     \vspace{-0.2cm}\item \texttt{pamela-security-patterns} is an add-on library containing some security-pattern implementations.
@@
 
-\subsection{On performance issues}
+\subsection{Performance analysis}
 
-Compared to a basic POJO implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpreted nature for some model features.
+Compared to a basic POJO implementation, the use of PAMELA implies a CPU and memory-footprint overhead.
+This is  due to the partially-interpreted nature of some of its features.
+We have developed a performance workbench in order to measure it \footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}}.
 
-A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developed to help measure that overhead.
-We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value).
+We first defined a base model composed of four entities declaring four to five properties (single and multi-valued) each.
 From this simple model, two Java implementations have been derived using code generation.
-The first implementation was made using fully implemented Java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).
+The first implementation uses fully implemented Java classes (POJO), while the second one is fully interpreted (it uses the PAMELA framework and defines only interfaces and API methods).
+Then, for each implementation, we have instantiated the base model (by creating 1010101 objects with their properties initialized to default values) and carried out performance measures.
+As expected, we measured a CPU overhead for the fully interpreted implementation (from $\times$20).
+This is due to \texttt{MethodHandler} mechanism which is fully interpreted and contains many hooks.
+Therefore, it  cannot be compared to a fully compiled and optimized Java dynamic binding.
+We also measured a memory footprint overhead, but of a less important weight that the CPU overhead ($\times$5 factor).
 
-Unsurprisingly, we quantify a significant CPU overhead (from $\times$20).
-This is due to the fact that \texttt{MethodHandler} mechanism is fully interpreted and contains many hooks, and cannot be compared to a fully compiled and optimized Java dynamic binding.
-As well, memory footprint is increased from an
-$\times$5 factor.
-
-These conclusions must be moderated by the fact that this overhead only applies
-to a very small portion of the code (mostly property accessors methods).
+Note however that the measured overhead only applies to a very small portion of the code (mostly property accessors methods).
 A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer.
 Using PAMELA won’t increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case).
-As we will see in the next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.
-
-Moreover, comparison between the two implementations should be tempered by the fact that the PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.).
+We have used the Yourkit Java Profiler\footnote{https://www.yourkit.com/} on PAMELA based applications and, as expected, the time passed on accessors was negligible compared to the time used on methods implementing business logic.
+As we will see in the next section, PAMELA has been successfully applied to many industrial projects, with no performance issues reported.
+Moreover, the PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.) which out-weight the performance overhead.
 

### 2021-04-23 implementation.tex
Updates from Overleaf

@@
 
-Compared to a basic POJO implementation, the use of PAMELA implies a CPU and memory-footprint overhead.
-This is  due to the partially-interpreted nature of some of its features.
+Compared to a basic POJO implementation, the use of PAMELA implies a CPU and memory footprint overhead.
+This is  due to the partially interpreted nature of some of its features.
 We have developed a performance workbench in order to measure it \footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}}.
@@
 
-Note however that the measured overhead only applies to a very small portion of the code (mostly property accessors methods).
+Note, however that the measured overhead only applies to a very small portion of the code (mostly property accessors methods).
 A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer.
@@
 As we will see in the next section, PAMELA has been successfully applied to many industrial projects, with no performance issues reported.
-Moreover, the PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.) which out-weight the performance overhead.
+Moreover, the PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc.) which outweigh the performance overhead.
 

### 2022-03-10 experimentations.tex
tex names standardization

--- 2021-03-23 6-Experimentations.tex
@@
 The full Openflexo infrastructure is composed of about 50 components.
-In most components, the PAMELA framework is extensively used.
+In most components, the PAMELA framework is largely used.
 As an example the \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagraming features) is composed of 998 classes.
@@
 The PAMELA implementation is now really stable and mature.
-According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no PAMELA-specific issue, while co-evolution of models and code is greatly improved compared to a "code generation"-based solution.
+According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no PAMELA-specific issue, while co-evolution of models and code is greatly improved compared to a 'code generation'-based solution.
 

### 2022-03-10 main.tex
tex names standardization
Updates from Overleaf

--- 2021-04-21 1-Main.tex
@@
 
-\input{2-Abstract}
+\input{abstract}
 
@@
 %% information and builds the first part of the formatted document.
+\clearpage
+\section*{Software and Code Metadata}
+\label{}
+
+\section*{Current executable software version}
+\label{}
+
+\begin{table}[!h]
+\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
+\hline
+\textbf{Nr.} & \textbf{(executable) Software metadata description} &  \\
+\hline
+S1 & Current software version & 1.5 \\
+\hline
+S2 & Permanent link to executables of this version  & https://maven.openflexo.org/artifactory/
+libs-release/org/openflexo/pamela-core/1.5 \\
+\hline
+S3 & Legal Software License & Dual licencing: GPLV3 and EUPL \\
+\hline
+S4 & Computing platform/Operating System & Java runtime environment \\
+\hline
+S5 & Installation requirements \& dependencies & JRE \geq 1.6\\
+\hline
+S6 & If available, link to user manual - if formally published include a reference to the publication in the reference list & https://pamela.openflexo.org\\
+\hline
+S7 & Support email for questions & contact@openflexo.org\\
+\hline
+\end{tabular}
+\caption{Software metadata}
+\label{} 
+\end{table}
+
+\section*{Current code version}
+\label{}
+
+\begin{table}[!h]
+\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
+\hline
+\textbf{Nr.} & \textbf{Code metadata description} &  \\
+\hline
+C1 & Current code version & 1.6.1 \\
+\hline
+C2 & Permanent link to code/repository used of this code version & https://github.com/openflexo-team/pamela \\
+\hline
+C3 & Legal Code License   & Dual licencing: GPLV3 and EUPL \\
+\hline
+C4 & Code versioning system used & git / github \\
+\hline
+C5 & Software code languages, tools, and services used & Java \\
+\hline
+C6 & Compilation requirements, operating environments \& dependencies & JDK \geq 1.6\\
+\hline
+C7 & If available Link to developer documentation/manual & https://pamela.openflexo.org \\
+\hline
+C8 & Support email for questions & contact@openflexo.org\\
+\hline
+\end{tabular}
+\caption{Code metadata}
+\label{} 
+\end{table}
 
@@
 
-%\section{Industrial experimentation and validation}
-%\label{sec:validation}
-%\input{validation.tex}
-
-\input{6-Experimentations}
+\input{experimentations}
 \label{sec:Experimentations}
@@
 
-%\input{4-ApproachOverview}
-
-%\input{5-Features}
-
-%\input{6-Experimentations}
-
-%\input{7-Relatedworks}
-
-\input{8-Conclusion}
+\input{conclusion}
 
